### 2019-05-20 paper.tex
criacao dos arquivos basicos

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{subfig}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Visual Analytics of Large Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Alan Valejo \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, BR\\
\email{renato.fabbri@gmail.com},
\email{alanvalejo@gmail.com},
\email{cristina@icmc.usp.br}\\
\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.

\keywords{Network visualization \and Multilevel strategies \and Visual analytics \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
%
%
\section{Introduction}
\subsection{Related work}
\subsection{Nomenclature remarks}
\section{A description of the method}\label{sec:des}
\begin{figure}[!h]\centering
% \includegraphics[width=.7\textwidth]{}
  \caption{.
  }\label{fig:glob}
\end{figure}

\noindent 
\section{Software implementation}\label{sec:imp}
\section{Conclusions and further work}\label{sec:con}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{biblio}
%
% \begin{thebibliography}{8}
% \bibitem{ref_article1}
% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
% 
% \bibitem{ref_lncs1}
% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
% Springer, Heidelberg (2016). \doi{10.10007/1234567890}
% 
% \bibitem{ref_book1}
% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
% Location (1999)
% 
% \bibitem{ref_proc1}
% Author, A.-B.: Contribution title. In: 9th International Proceedings
% on Proceedings, pp. 1--2. Publisher, Location (2010)
% 
% \bibitem{ref_url1}
% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
% Oct 2017
% \end{thebibliography}
\end{document}

### 2019-05-20 paper.tex
resumo e anotacoes

@@
 %
-\title{Visual Analytics of Large Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
+\title{Visual Analytics of Large Bipartite Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
 %
-%\titlerunning{Abbreviated paper title}
+\titlerunning{Multilevel Visualization of Bipartite Networks}
 % If the paper title is too long for the running head, you can set
@@
 %
-\maketitle              % typeset the header of the contribution
+\maketitle
 %
@@
 % 150--250 words.
+  It is a well established fact that bipartite networks are often used to
+  model real-world phenomena and that it plays fundamental roles in
+  graph theory.
+  Multilevel strategies have been developed for optimization tasks, and for the visualization of simple (``unipartite'') networks, but not for bipartite networks.
+  In this work, we present advances in the use of multilevel strategies for the
+  visualization of bipartite networks, allowing interactive and intuitive navigation of such structures and visual mappings of large datasets.
+  More specifically, we developed a visual analytics web interface in which
+  bipartite networks are presented to the user, by a parametrizable simpification
+  of the original structures, through the application of coarsening algorithms.
+  The resulting networks are then presented to the user,
+  providing a genuine route for the ``overview first - focus on demand''
+  process on the analysis of the undelying data, in which the analyst
+  selects supervertices or whole network sectors for more detailed observation,
+  i.e. performs requests for the interface to display 
+  specific structures in less simplified settings.
+  Moreover, the application is useful for the development multilevel strategies
+  e.g. by the specification of vertices to guide the coarsening processes and
+  the examination of resulting multilevel hierarchy.
 
@@
 \subsection{Related work}
-\subsection{Nomenclature remarks}
-\section{A description of the method}\label{sec:des}
+%%%
+% artigos que devemos citar pelo histórico científico
+% da estratégia ML, da visualização ML
+% e os que devemos citar pelo ML em redes bipartidas
+% e desenvolvidos no ICMC
+
+\subsection{Nomenclature and conceptual remarks}
+%%%
+% level vs layer
+% simple and heterogeneous graphs
+% multiplex, ?
+% multilevel strategies for optimization
+% supernode metanode,
+
+\section{Method description}\label{sec:des}
+%%%
+% mlpb and other methods developed
+% arguments for the preferred method (mlpb?)
 \begin{figure}[!h]\centering
@@
 \section{Software implementation}\label{sec:imp}
+%%%
+% description of the functionalities/tools implemented,
+% of the technologies and libraries used
+% optimization of the computational capacity for large networks by
+% means of Web GL, triangles for nodes and details on demand
+
+\section{Results and discussion}
+%%%
+% consistent use of overview first, details on demand
+% a first contribution on ML vis. of bipartite networks
+% simple interactivity strategies
+
 \section{Conclusions and further work}\label{sec:con}
+%%%
+% extension to multipartite (or heterogeneous) networks.
+% use for marking nodes as pivots for coarsening
 
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
 \bibliographystyle{splncs04}

### 2019-05-22 paper.tex
adicionada bibliografia principal e escrita boa parte da intro

@@
 % 150--250 words.
-  It is a well established fact that bipartite networks are often used to
-  model real-world phenomena and that it plays fundamental roles in
+  It is a well established fact that bipartite, or two-layer, networks are pervasive
+  in model real-world phenomena and that they play fundamental roles in
   graph theory.
-  Multilevel strategies have been developed for optimization tasks, and for the visualization of simple (``unipartite'') networks, but not for bipartite networks.
+  Multilevel strategies have been developed for optimization tasks,
+  and for the visualization of simple (``unipartite'') networks,
+  but their employment for visualizing bipartite networks were not found by the authors.
   In this work, we present advances in the use of multilevel strategies for the
-  visualization of bipartite networks, allowing interactive and intuitive navigation of such structures and visual mappings of large datasets.
+  visualization of bipartite networks,
+  allowing interactive and intuitive navigation of such structures and visual mappings of large datasets.
   More specifically, we developed a visual analytics web interface in which
-  bipartite networks are presented to the user, by a parametrizable simpification
-  of the original structures, through the application of coarsening algorithms.
+  a parametrizable simpification of
+  bipartite networks are obtained through the application of coarsening algorithms.
   The resulting networks are then presented to the user,
@@
   e.g. by the specification of vertices to guide the coarsening processes and
-  the examination of resulting multilevel hierarchy.
+  the examination of the resulting multilevel hierarchy.
 
@@
 %
-%
-%
 \section{Introduction}
-\subsection{Related work}
+The visualization of large-scale networks poses challenges both in terms of computational costs
+and of effective presentation of the information for the user~\cite{tang,staudt}.
+These issues may be aggravated in the case of bipartite networks,
+due to their sparsity and topological complexities~\cite{alan1}.
+Bipartite networks are comprised of two partitions of nodes,
+called ``layers'',
+and links are not incident between nodes in the same partition.
+Such network type arises very often and naturally from the representation
+of relations among two kinds objects,
+e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
+Furthermore, real-world networks are often bipartite, and most unipartite networks
+are projections of bipartite networks or may be considered as exibiting bipartite properties~\cite{guillaume0,guillaume}.
+In order to assist the visualization and navigation of large networks, one possibility is the use of
+multilevel strategies, which consist on the employment of incremental coarsening of the original
+network to obtain a sequence of simplified representations.
+Multilevel strategies are most traditionally used for executing complex algorithms
+on large-scale networks by the application of the algorithm on a smaller
+version of the network~\cite{alan2,ml2}.
+Their employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
+networks have been reported, but their exploitation for visualizing bipartite networks
+was not found by the authors.
+Accordingly, we present a system for the visualization of bipartite networks using
+multilevel strategies developed for bipartite networks.
+The system consists on presented a simplified version of the network for the user, which then
+requests for supervertices (or collections of them) to be uncoarsened and presented in more detail.
+
+This paper is organized as follows: in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
+Fundamental concepts are introduced in Section~\ref{bac}, and the method is delineated in Section~\ref{des}.
+The software implementation is then described in Section~\ref{sof}.
+Results and discussion are in~\ref{res}.
+Finally, Section~\ref{con} holds concluding and further work statements.
+
+\subsection{Related work}\label{rel}
 %%%
@@
 % e desenvolvidos no ICMC
+Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
+Also, the aggregation of clusters have been reported, and comprises an approach that resembles
+the coarsening procedure in creating simplified representations of the original network~\cite{a1,a2,a3,a4,a5,a6}.
+Even so, the authors are not aware of previous reports on the use of multilevel strategies
+for the visualization and navigation of bipartite networks.
 
-\subsection{Nomenclature and conceptual remarks}
+\subsection{Nomenclature and conceptual remarks}\label{nom}
 %%%
@@
 
-\section{Method description}\label{sec:des}
+\section{Backgroud}\label{bac}
+\section{Method description}\label{des}
 %%%
@@
 \noindent 
-\section{Software implementation}\label{sec:imp}
+\section{Software implementation}\label{sof}
 %%%
@@
 
-\section{Results and discussion}
+\section{Results and discussion}\label{res}
 %%%
@@
 
-\section{Conclusions and further work}\label{sec:con}
+\section{Conclusions and further work}\label{con}
 %%%

### 2019-05-22 paper.tex
feito ate os conceitos fundamentais

@@
   More specifically, we developed a visual analytics web interface in which
-  a parametrizable simpification of
+  a parametrizable simplification of
   bipartite networks are obtained through the application of coarsening algorithms.
@@
   providing a genuine route for the ``overview first - focus on demand''
-  process on the analysis of the undelying data, in which the analyst
+  process on the analysis of the underlying data, in which the analyst
   selects supervertices or whole network sectors for more detailed observation,
@@
 Furthermore, real-world networks are often bipartite, and most unipartite networks
-are projections of bipartite networks or may be considered as exibiting bipartite properties~\cite{guillaume0,guillaume}.
+are projections of bipartite networks or may be considered as exhibiting bipartite properties~\cite{guillaume0,guillaume}.
 In order to assist the visualization and navigation of large networks, one possibility is the use of
@@
 This paper is organized as follows: in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
-Fundamental concepts are introduced in Section~\ref{bac}, and the method is delineated in Section~\ref{des}.
-The software implementation is then described in Section~\ref{sof}.
+The method is delineated in Section~\ref{des} and
+the software implementation is then described in Section~\ref{sof}.
 Results and discussion are in~\ref{res}.
@@
 for the visualization and navigation of bipartite networks.
+Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system for visual analytics, i.e. the other works do not
+address data analysis by interactive visual interfaces,
+including all multilevel strategies for visualization.
 
-\subsection{Nomenclature and conceptual remarks}\label{nom}
+\subsection{Nomenclature remarks}\label{nom}
 %%%
@@
 % supernode metanode, 
+The main vocabulary issue that arises in the context of this article is between
+level and layer. Layers are the sets of a bipartite network, while levels are the sequence
+of coarsened networks.
+Also, simple or homogeneous networks are opposed to heterogeneous networks, in which there
+are more then one type of node.
+Bipartite networks may be regarded as an elementary case of heterogeneous networks,
+but it has one additional restriction: only nodes of different types are connected.
+Another usual concept in this context is that of a multilayer network, in which there are layers,
+i.e. partitions of nodes, in addition to nodes and links.
+Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layers.
+Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
+A thorough exposition of the vocabulary are beyond the scope of this article, but
+some attention for the issue is helpful to assist searches and newcomers.
 
-\section{Backgroud}\label{bac}
 \section{Method description}\label{des}
@@
 % arguments for the preferred method (mlpb?)
+\subsection{Fundamental concepts}\label{bac}
+A bipartite network $G=(V,E)$ consists in a set $V$ of vertices which is partitioned in two subsets
+with no links between nodes in the same set, i.e. $\exists\; V_1, V_2:\; V_1\cup V_2 = V,\;V_1\cap V_2 = \emptyset$, and $E\subseteq V_1 \times V_2$.
+The variable names are borrowed from more traditional Graph theory, with Vertices (nodes) and Edges (links).
+
+A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network.
+The coarsening procedures requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
+from the matched nodes.
+There are several coarsening algorithms reported in the scientific literature~\cite{}, a few of them
+developed for bipartite networks~\cite{}.
+A discussion of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
+Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks, but because it has yielded simplified networks which carries essential topological features of the original networks, and that result is perceived is the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.
+
+Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{}.
+Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
+Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the infererence of relationships, and providing a manipulable medium for the exploration of the information of interest~\cite{}.
+
+\subsection{The navigation pathway}
+
 \begin{figure}[!h]\centering

### 2019-05-22 paper.tex
feito ate os conceitos fundamentais

@@
 Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
-Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the infererence of relationships, and providing a manipulable medium for the exploration of the information of interest~\cite{}.
+Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships, and providing a manipulable medium for the exploration of the information of interest~\cite{}.
 

### 2019-05-22 paper.tex
retoques de escrita

@@
 The visualization of large-scale networks poses challenges both in terms of computational costs
-and of effective presentation of the information for the user~\cite{tang,staudt}.
+and of effective presentation of the information to the user~\cite{tang,staudt}.
 These issues may be aggravated in the case of bipartite networks,
@@
 Furthermore, real-world networks are often bipartite, and most unipartite networks
-are projections of bipartite networks or may be considered as exhibiting bipartite properties~\cite{guillaume0,guillaume}.
+are projections of bipartite networks or exhibit bipartite properties~\cite{guillaume0,guillaume}.
 In order to assist the visualization and navigation of large networks, one possibility is the use of
@@
 version of the network~\cite{alan2,ml2}.
-Their employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
+The employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
 networks have been reported, but their exploitation for visualizing bipartite networks
-was not found by the authors.
+was not found by the authors, as described in the next subsection.
 Accordingly, we present a system for the visualization of bipartite networks using
 multilevel strategies developed for bipartite networks.
-The system consists on presented a simplified version of the network for the user, which then
-requests for supervertices (or collections of them) to be uncoarsened and presented in more detail.
+The system consists on presenting a simplified version of the network to the user, which then
+requests for supervertices (or collections of them) to be uncoarsened and visualized in more detail.
 
-This paper is organized as follows: in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
+This paper is organized as follows: in Section~\ref{rel} the related work is examined,
+while in~\ref{nom} are selected remarks about the vocabulary.
 The method is delineated in Section~\ref{des} and
@@
 Results and discussion are in~\ref{res}.
-Finally, Section~\ref{con} holds concluding and further work statements.
+Finally, Section~\ref{con} holds concluions and further work envisioned.
 
@@
 for the visualization and navigation of bipartite networks.
-Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system for visual analytics, i.e. the other works do not
-address data analysis by interactive visual interfaces,
-including all multilevel strategies for visualization.
+Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system suitable for visual analytics, i.e. the other works do not
+address data analysis by interactive visual interfaces, including all multilevel strategies for visualization.
 
@@
 The main vocabulary issue that arises in the context of this article is between
-level and layer. Layers are the sets of a bipartite network, while levels are the sequence
+level and layer. Layers are the (two) node partitions of a bipartite network, while levels comprise the sequence
 of coarsened networks.
@@
 Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
-A thorough exposition of the vocabulary are beyond the scope of this article, but
+A thorough exposition of the vocabulary is beyond the scope of this article, but
 some attention for the issue is helpful to assist searches and newcomers.
@@
 
-A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network.
+A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network
+and where $|V_i| \leq |V_{i+1}|$.
 The coarsening procedures requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
@@
 A discussion of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
-Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks, but because it has yielded simplified networks which carries essential topological features of the original networks, and that result is perceived is the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.
+Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks,
+but because it has yielded simplified networks which carries essential topological features of the original networks,
+and that result is perceived in the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.
 
@@
 Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
-Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships, and providing a manipulable medium for the exploration of the information of interest~\cite{}.
+Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways,
+which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships,
+and providing a manipulable medium for the exploration of the information of interest~\cite{}.
 

### 2019-05-23 paper.tex
alcancada secao de implementacao em software

@@
 \usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{algorithmicx}
+\usepackage{algorithm2e}
 \usepackage[utf8]{inputenc}
@@
   visualization of bipartite networks,
-  allowing interactive and intuitive navigation of such structures and visual mappings of large datasets.
+  allowing for interactive and intuitive navigation of such structures and visual mappings of large datasets.
   More specifically, we developed a visual analytics web interface in which
@@
   specific structures in less simplified settings.
-  Moreover, the application is useful for the development multilevel strategies
+  Moreover, the application is useful for the further development multilevel strategies themselves
   e.g. by the specification of vertices to guide the coarsening processes and
@@
 with no links between nodes in the same set, i.e. $\exists\; V_1, V_2:\; V_1\cup V_2 = V,\;V_1\cap V_2 = \emptyset$, and $E\subseteq V_1 \times V_2$.
-The variable names are borrowed from more traditional Graph theory, with Vertices (nodes) and Edges (links).
+The variable names are borrowed from more traditional
+Graph (network) theory, with Vertices (nodes) and Edges (links).
+One may regard the network as $G=(V,E,\sigma,\omega)$,
+$\sigma\,:\,V\rightarrow \mathbb{R}^*$ and with $\omega\,:\,E \rightarrow \mathbb{R}^*$,
+where $\sigma(v)$ is the weight of the node $v$ and
+$\omega(u,v)$ is the weight of the link $(u,v)$.
 
@@
 developed for bipartite networks~\cite{}.
-A discussion of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
-Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks,
-but because it has yielded simplified networks which carries essential topological features of the original networks,
-and that result is perceived in the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.
+An exposition about these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
+Most importantly, we are interested in coarsening suited for bipartite networks,
+as it has yielded simplified networks which present essential topological features
+of the original networks,
+and such result is perceived in the visualization as hinted in\~cite{alan2} and further shown in
+Sections~\ref{des},~\ref{sof} and~\ref{res}.
 
@@
 
-\subsection{The navigation pathway}
+\subsection{Multilevel coarsening of bipartite networks}\label{bac}
+Before~\cite{alan2}, multilevel approaches were not directly aplicable to bipartite networks.
+Such work introduced novel and efficient matching and coarsening algorithms,
+and scrutinized their validity for solving optimization problems, dimensionality reduction,
+and in the preservation of essential topological features of the original network.
+We here describe the outline of such procedures in the context of our visual analytics contribution.
+
+The multilevel optimization is a metaheuristics used to guide,
+modify and potentially fix a solution obtained from a target algorithm.
+It is divided into three phases: coarsening, solution finding and uncoarsening,
+where the solution found in the coarsest network $G_L$ is projected back to the original,
+uncoarsened, network $G_0$.
+Figure~\ref{mlf} illustrates such process.
 
 \begin{figure}[!h]\centering
-% \includegraphics[width=.7\textwidth]{}
-  \caption{.
-  }\label{fig:glob}
+ \includegraphics[width=\textwidth]{mlf}
+  \caption{Phases of the multilevel optimization.
+  A sequence of incrementally reduced networks is obtained in the coarsening phase.
+  An initial solution found at the coarsest network, in the solution finding phase,
+  is projected back to the initial, uncoarsened, network, in the uncoarsening phase.
+  Further details are given in Section~\ref{bac} and specificities related to these
+  phases when applied to visual analytics are in Sections~\ref{net} and~\ref{nav}.
+  }\label{mlf}
 \end{figure}
+
+The coarsening phase results in a sequence of networks $G_l$ from the initial network $G_0$,
+on multiple levels of details.
+The process is carried out by two algorithms: \emph{matching}, which decides which nodes to merge,
+and \emph{contracting}, which achieves the reduced representation from the network and the matching.
+In general, pairs of nodes are selected to be merged into supernodes, and
+most often a matching $M$ consists in a set of non-adjacent links,
+i.e. $\forall\; l_1, l_2 \in M, u \in l_1 \Rightarrow u \notin l_2$.
+The heavy-edge mathing, for example, is a matching that fits this canonical description
+in attempting to maximize total matching weight, but the matching may not satisfy such restrictions.
+Selecting clicks and other larger node sets to be merged into supernodes are possibilites
+under development.
+%%%
+% citar alguem aqui?
+For bipartite network, we use the algorithms provided by~\cite{alan2} in which two restrictions
+are imposed to the matching:
+\begin{itemize}
+  \item A node may only match nodes on the same layer.
+  \item A node may only match nodes that are reachable by two successive links
+    (i.e. the closest possible nodes on the same layer).
+\end{itemize}
+
+The matching is followed by the contraction of the network into a coarser form.
+Typically, the nodes matched are merged into a supernode
+with the weight equal to the sum of the weight of nodes it contains,
+and the links incident to such nodes are joined into superlinks among the supervertices,
+also with total weight equal to the sum of the weight of the links merged.
+The solution finding is usually reasoned about in terms of the computational cost, being much
+alleviated in the reduced network.
+The uncoarsening is usually performed through the complete hierarchy,
+from the coarsest to the original network, with successive refinement of the solutions to
+avoid local minima and improve solution quality.
+These two phases are substantially different when the multilevel strategy is applied to visualization,
+which motivated their separate exposition in the next subsection.
+
+\subsection{Network visualization assisted by multilevel strategies}\label{net}
+In applying the multilevel framework to visualization,
+the target algorithm is the visual mapping of the network,
+i.e. the solution finding phase is the visualization of the network by means of
+the reduced representation.
+The uncoarsening is performed only on demand, through user requests
+to visualize network regions with more detail, thus avoiding unecessary complexity and,
+most importantly, avoiding to overload the visualization with information beyond the
+cognitive convenience of the user and the computational power that the machine being used
+is able to provide for real-time interactivity.
+
+The interactivity is crucial for a number of reasons:
+the definition of the uncoarsening phase desired;
+for the navigation of the network,
+including the access to metadata and changes to the visualization achieved;
+for uncoarsening supernodes;
+and for tunning the achieved visual mapping status, such as by resizing
+and moving nodes.
+The overall procedure is delineated in Algorithm~\ref{alg} and further
+detailed in Sections~\ref{nav} and~\ref{sof}.
+
+\SetKwData{Matching}{Matching$_b$}
+\SetKwData{Contracting}{Contracting$_b$}
+\SetKwData{map}{map\_to\_screen}
+\SetKwData{trans}{transform\_visual\_mapping}
+\SetKwInOut{Return}{Return}
+\begin{algorithm}
+  \KwIn{\\
+  \quad bipartite network: $G$\\
+  \quad maximum number of levels: $L \in [0,n] \subset \mathbb{Z}$\\
+  \quad reduction factor: $rf \in (0,0.5] \subset \mathbb{R}$.\\
+  \quad layers to be coarsened: $layers \in {1,2}$\\
+  \quad user command given through the visual interface: $C$
+  }
+  \KwOut{\\
+  \quad Visual mapping of the network: $V$}
+  \BlankLine
+  $i \leftarrow 1$\;
+  \While{$i \le$ layers}{
+    $l\leftarrow 1$\;
+    \While{$l\leq L$}{
+      $M \leftarrow$ \Matching{$G_l$, $i$, $rf$}\;
+      $G_{l+1} \leftarrow$ \Contracting{$G_l$, $M$}\;
+      increment $l$\;
+    }
+      increment $i$\;
+  }
+  $V \leftarrow$ \map{$G_L$}\;
+  \While{1}{
+    $C\leftarrow$ user command\;
+    $V\leftarrow$ \trans{$V$, $C$}\;
+  }
+  \caption{An algorithmic description of the multilevel strategy adapted for the visualization of bipartite networks.
+  The routines {\bf Matching$_b$} and {\bf Contracting$_b$} are any of the matching or contracting
+  algorithms suited for bipartite networks, such as described in~\cite{alan2} which are only ones
+  currently available to the knowledge of the authors.
+  The {\bf map\_to\_screen} routine is performed through the use of a network layout algorithm and
+  then the rendering of the network to the screen.
+  The visual mapping may then be transformed by the user by requesting uncoarsening of specific supernodes,
+  or by other commands not specific to the multilevel strategy, such as requesting metadata exposition, changes to the position of nodes, the color or transparency of nodes and links, zoom and pan,
+  or any other operation defined in Section~\ref{sof}.}\label{alg}
+\end{algorithm}
+
+\subsection{The navigation pathway}\label{nav}
+The abundance of information within large networks makes pertinent the
+application of
+the ``visual information-seeking mantra'', also known as Shneiderman's mantra:
+\emph{overview first, zoom and filter, then details-on-demand}.
+This mantra comprises a number of visual design guidelines,
+such as the details-on-demand technique, and provides
+a very acknowledged framework for designing information visualization applications.
+Accordingly, our idealized exploration pathway starts with an overview,
+achieved by the mapping the coarsest representation of the network through
+standard network layout algorithms for node-link diagrams.
+The user may then zoom into specific regions of the network, and then request
+details by a number of operations, most importantly:
+\begin{itemize}
+  \item reposition nodes, exposing linking patterns.
+  \item request metadata, such as number of children, the parent in a subsequent level, or the number of neighbors.
+  \item request visual mapping of network features, such as node size related to number of neighbors or the number of child nodes it represents.
+  \item request the uncoarsening of specific supernodes or any arbitrary group of supernodes.
+\end{itemize}
+
+Other operations are convenient to make the visual mapping adequate for the diverse settings possible:
+network size, open supervertices, levels exposed, and topological features such as commnity structures.
 

### 2019-05-23 paper.tex
adiantado ate soft implementation

@@
 for uncoarsening supernodes;
-and for tunning the achieved visual mapping status, such as by resizing
+and for tuning the achieved visual mapping status, such as by resizing
 and moving nodes.
@@
 Other operations are convenient to make the visual mapping adequate for the diverse settings possible:
-network size, open supervertices, levels exposed, and topological features such as commnity structures.
+network size, open supervertices, levels exposed, and topological features such as community structures.
+The navigations pathway, already as achieved within the software we made available, is illustrated in Figure~\ref{fnav}.
+
+\begin{figure}[!h]\centering
+ \includegraphics[width=0.6\textwidth]{fnav}
+  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
+  and auxiliary tools.
+  }\label{fnav}
+\end{figure}
 
 \noindent 
-\section{Software implementation}\label{sof}
+\section{Software implementation: mlvis}\label{sof}
 %%%
@@
 % means of Web GL, triangles for nodes and details on demand
+We implemented the framework described in Section~\ref{des} using scientific,
+database and web resources,
+and made the final result available to the user within a web page
+for use through simple mouse-driven actions
+and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
+The final result is exemplified in Figure~\ref{fpage} and next subsections
+describe its functionalities and the technologies used.
+
+\begin{figure}[!h]\centering
+ \includegraphics[width=\textwidth]{fpage}
+  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
+  and auxiliary tools.
+  }\label{fpage}
+\end{figure}
+
+\subsection{The use of mlvis}
+% network upload, format.
+% layouts available
+% algorithms available for coarsening, and their parametrization
+% initialization of the visualization: click on render
+% tools available and their operation
+
+\subsection{Technologies used within mlvis}
+% coarsening algorithms
+% flask for server, feather for auxiliary server
+% mongodb filesystem for data
+% Nuxt/Vue.js for the frontend with heavy use of PIXI.js
 
@@
 
+% enhancements of the navigation through the use of being-developed coarsening algorithms
+% the use of the interface for guiding coarsening (through markers)
+% the use of the interface to analyse coarsening results
+
+% technologies used: maybe simpler if using meteor.js
+% and simpler to achieve multiuser interactivity
+
 \section{Conclusions and further work}\label{con}
@@
 % extension to multipartite (or heterogeneous) networks.
-% use for marking nodes as pivots for coarsening
 

### 2019-05-24 paper.tex
adiantado ate soft implementation
terminado perfil geral do escrito

@@
 the software implementation is then described in Section~\ref{sof}.
-Results and discussion are in~\ref{res}.
+% Results and discussion are in~\ref{res}.
 Finally, Section~\ref{con} holds concluions and further work envisioned.
@@
 and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
-The final result is exemplified in Figure~\ref{fpage} and next subsections
+The final result is exemplified in Figures~\ref{fpage0} and~\ref{fpage} and next subsections
 describe its functionalities and the technologies used.
+
+\begin{figure}[!h]\centering
+ \includegraphics[width=\textwidth]{fpage0}
+  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
+  and auxiliary tools.
+  }\label{fpage0}
+\end{figure}
 
@@
 % tools available and their operation
+The exposition on this section is suported by Figure~\ref{usage}.
+The analyst first selects a network of interest to be explored.
+One may also upload a network and then select it on the same drop-down menu.
+The software is then ready to render the network on the screen for arbitrary
+navigation, but the user may wish to select one of the available network layout
+algorithms and tune the coarsening to be performed.
+Upon request given by hiting the ``render network'' button,
+the network is mapped to the screen.
+The subsequent usage relies on the manipulation of the visual mapping presented on the
+canvas by means of actions
+performed through managing the tools and related mouse operations.
 
-\subsection{Technologies used within mlvis}
+There are multiple levels of representation, which is entailed by the multilevel strategy.
+This makes pertinent that actions be performed in only one of the available levels.
+Thus, there is always a ``level on focus'' in which the tools perform actions, such
+as link coloring or opening supernodes to expose their child nodes.
+The first button on the toolbar is thus for changing the level on focus.
+The second set of buttons are for changing how the nodes are visualized:
+color, size, if size is related to degree or number of children, restore node size,
+node transparency, and rotation.
+The third set of tool buttons if for links: color, thickness, transparency, if thickness and
+transparency are related to link weight, and restor link properties.
+
+The fourth set of tool buttons is more specific to the needs of the MlBiNetViz, and
+are selected to make mouse operations entail special processes and transformations of the visualization.
+The first button is selected for the interface to return core information from a clicked node:
+id, number of children, id of parent, number of neighbors (degree), sum of weight of links (strength), and other information the analysts report necessity of.
+The second tool button is seleted to move nodes: specific nodes are individually moved if clicked and dragged; a rectangular region is made available for moving all nodes therein if an empty region of the canvas is clicked and the mouse cursor dragged before releasing the button.
+The third tool button is selected for opening a supernode and exposing their child nodes and links among them:
+a region is defined by the user by clicking on an empty space on the canvas and gragging the mouse before releasing, the supernodes therein are joined into a rectangular region that contains all of them, and their child nodes (on a finer level of detail) are drawn inside such rectangular region.
+Open supernodes may be joined using the same tool, but it was found convenient to make available a
+fourth button to assist in this task: if this tool is selected, one may join two open supernodes by
+clicking on them in sequence.
+The last tool button is selected to perform three operations on open supernodes:
+if an open supernode is clicked on and dragged, the rectangular shape is resized accordingly;
+if there is no drag, i.e. the super node is clicked on and released, the links from it to the other
+supernodes are attached to the next counterclockwise corner of its rectagular region.
+When the open supernode is resized, the layout of its child nodes is optimized for the drawn region.
+This is most useful when used in combination with the drag tool: on the finer level of detail, the nodes are repositioned into a more compact disposition. Then upon a (potentially minimal) supernode
+risize, the child nodes are made to fit the whole rectangular area.
+
+The last set of tools acts on the whole canvas: they are dedicated to zoom and left-right, up-down pan, changing the color of the background, and toggling between current zoom and pan status and the initial setting.
+
+There are many subleties on the orchestrated use of these tools.
+As a facilitated means to convey the navigation possibilities and
+procedures available, the user is invited to watch a demo video in~\cite{yvideo},
+and use the MlBiNetViz in the instance we made availabel~\cite{mbnvpage}.
+The more seasoned software developer may browser the code, upload a local instance
+of the software, and make changes to its source code~\cite{mbnvcode}.
+
+\subsection{Software technologies employed within mlvis}
 % coarsening algorithms
@@
 % Nuxt/Vue.js for the frontend with heavy use of PIXI.js
+Most important parts of MlBiNetViz are written in a combination of JavaScript and Python:
+Vue.js (set up by Nuxt.js) is used to assist in the client (frontend),
+the server (backend) is a Flask Python server, used to perform
+specific or heavy calculations.
+There is a secondery server, a node FeatherJS, used to facilitate contact with the database
+and real-time multi-user interaction.
+The data is stored in a MongoDB database and ordinarily in the filesystem,
+while the multi-user interaction was deactivated for now to avoid unecessary complexity.
+The coarsening algorithms for bipartite networks are available in~\cite{bialgs}
+and are used by the Flask server.
+The fast WebGL 2D rendering on the canvas is performed using Pixi.js.
+This is a very brief account of the styles and software resources used to implement
+MlBiNetViz. The interested reader is welcome to browse the software code and documentation
+and contact the authors~\cite{mbvncode}.
 
-\section{Results and discussion}\label{res}
+% \section{Results and discussion}\label{res}
 %%%
@@
 % simple interactivity strategies
-
 % enhancements of the navigation through the use of being-developed coarsening algorithms
@@
 % extension to multipartite (or heterogeneous) networks.
+The resulting conceptual organization is very consistent with the Shneiderman's mantra,
+as exemplified the by the implemented software.
+Also, it is the first contribution to bipartite networks visual analytics
+assited by multilevel strategies.
+The interface developed is efficiently managed through simple interactive operations.
+
+The navigation may be enhanced by the use of in-development coarsening algorithms.
+The visualization pathway should be used to assist multilevel strategies by enabling the user
+to select nodes as pivots to guide the coarsening procedure.
+Also, the interface is envioned to assist the analysis of coarsening results.
+The bipartite framework is naturally extended to heterogeneous (or multipartite) networks,
+which may be a research route we will follow if there is enough interest of the researchers involved.
 

### 2019-05-28 paper.tex
terminada revisao do texto, falta so colocar as figuras e citacoes faltantes

--- 2019-05-23 paper.tex
@@
 %
-\title{Visual Analytics of Large Bipartite Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
+\title{Interactive Visualization of Large Bipartite Networks Assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
 %
@@
   It is a well established fact that bipartite, or two-layer, networks are pervasive
-  in model real-world phenomena and that they play fundamental roles in
+  in modeling real-world phenomena and that they play fundamental roles in
   graph theory.
@@
   allowing for interactive and intuitive navigation of such structures and visual mappings of large datasets.
-  More specifically, we developed a visual analytics web interface in which
+  More specifically, we developed a web-based visualization interface in which
   a parametrizable simplification of
@@
   specific structures in less simplified settings.
-  Moreover, the application is useful for the further development multilevel strategies themselves
+  The application allows not only the visualization of large networks,
+  but also for an on-demand navigation of the multilevel structure,
+  and is useful for the further development multilevel strategies themselves
   e.g. by the specification of vertices to guide the coarsening processes and
@@
 
-\keywords{Network visualization \and Multilevel strategies \and Visual analytics \and Big data \and Complex networks \and Data visualization.}
+\keywords{Network visualization \and Bipartite networks \and Multilevel strategies \and Big data \and Complex networks \and Data visualization.}
 \end{abstract}
@@
 multilevel strategies, which consist on the employment of incremental coarsening of the original
-network to obtain a sequence of simplified representations.
+network to obtain a sequence of simplified representations, i.e. with fewer nodes and links.
 Multilevel strategies are most traditionally used for executing complex algorithms
@@
 The employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
-networks have been reported, but their exploitation for visualizing bipartite networks
+networks has been reported, but their exploitation for visualizing bipartite networks
 was not found by the authors, as described in the next subsection.
@@
 requests for supervertices (or collections of them) to be uncoarsened and visualized in more detail.
+Thus, the resulting interface allows for the visualization of large structures;
+enables an interactive navigation of the structure e.g. by uncoarsening supernodes, zooming and requesting information about the nodes;
+and may assist not only in the analysis of bipartite networks, but also on the analysis and development of multilevel algorithms in bipartite networks.
+Most importantly, the overall method, and the implemented interface, is a proof-of-concept, it demonstrates the feasibility of the use of multilevel strategies for the efficient visualization of large bipartite networks.
 
@@
 the software implementation is then described in Section~\ref{sof}.
-Results and discussion are in~\ref{res}.
-Finally, Section~\ref{con} holds concluions and further work envisioned.
+% Results and discussion are in~\ref{res}.
+Finally, Section~\ref{con} holds conclusions and further work envisioned.
 
@@
 %%%
-% artigos que devemos citar pelo histórico científico
-% da estratégia ML, da visualização ML
-% e os que devemos citar pelo ML em redes bipartidas
-% e desenvolvidos no ICMC
+% incrementar? no artigo do alan tem mais informação sistematizada
+% citar mais trabalhos desenvolvidos no ICMC?
+% tratar mais especificamente de visualização de redes bipartidas?
+% tratar mais especificamente de estratégias multinivel?
 Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
@@
 for the visualization and navigation of bipartite networks.
-Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system suitable for visual analytics, i.e. the other works do not
-address data analysis by interactive visual interfaces, including all multilevel strategies for visualization.
+Furthermore, only~\cite{a1,a2,a3,a4,a6} report on interactive systems for network visualization,
+and these works do not tackle the use of multilevel strategies for visualization.
 
@@
 %%%
-% level vs layer
-% simple and heterogeneous graphs
+% remover seção?
 % multiplex, ?
-% multilevel strategies for optimization
-% supernode metanode, 
+% multilevel strategies synonym for optimization algorithms?
 The main vocabulary issue that arises in the context of this article is between
@@
 i.e. partitions of nodes, in addition to nodes and links.
-Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layers.
+Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layer.
 Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
@@
 \section{Method description}\label{des}
-%%%
-% mlpb and other methods developed
-% arguments for the preferred method (mlpb?)
 \subsection{Fundamental concepts}\label{bac}
-A bipartite network $G=(V,E)$ consists in a set $V$ of vertices which is partitioned in two subsets
+A bipartite network $G=(V,E)$ consists in a set $V$ of nodes which is partitioned in two subsets
 with no links between nodes in the same set, i.e. $\exists\; V_1, V_2:\; V_1\cup V_2 = V,\;V_1\cap V_2 = \emptyset$, and $E\subseteq V_1 \times V_2$.
@@
 Graph (network) theory, with Vertices (nodes) and Edges (links).
-One may regard the network as $G=(V,E,\sigma,\omega)$,
-$\sigma\,:\,V\rightarrow \mathbb{R}^*$ and with $\omega\,:\,E \rightarrow \mathbb{R}^*$,
+One may regard the network as $G=(V,E,\sigma,\omega)$, with
+$\sigma\,:\,V\rightarrow \mathbb{R}^*$ and $\omega\,:\,E \rightarrow \mathbb{R}^*$,
 where $\sigma(v)$ is the weight of the node $v$ and
@@
 A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network
-and where $|V_i| \leq |V_{i+1}|$.
-The coarsening procedures requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
+and where $|V_i| \geq |V_{i+1}|$.
+The coarsening procedure requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
 from the matched nodes.
@@
 developed for bipartite networks~\cite{}.
-An exposition about these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
+A detailed exposition of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
 Most importantly, we are interested in coarsening suited for bipartite networks,
@@
 of the original networks,
-and such result is perceived in the visualization as hinted in\~cite{alan2} and further shown in
-Sections~\ref{des},~\ref{sof} and~\ref{res}.
+and such result is perceived in the visualization as hinted in~\cite{alan2} and further shown in
+Section~\ref{sof}.
 
@@
 and providing a manipulable medium for the exploration of the information of interest~\cite{}.
+For this present work to be better characterized as a visual analytics contribution, the methods described in this section and software described in Section~\ref{sof} should further enable the discovery of information, such as metadata of the networks (e.g.
+authors, words or documents).
+Accordingly, our methods and software may be understood as data visualization with explicit visual analytics elements, such as interactivity, on-demand supply of information, and selective navigation.
 
@@
 and in the preservation of essential topological features of the original network.
-We here describe the outline of such procedures in the context of our visual analytics contribution.
+We here delineate such procedures in the context of our network visualization contribution.
 
@@
   An initial solution found at the coarsest network, in the solution finding phase,
-  is projected back to the initial, uncoarsened, network, in the uncoarsening phase.
+  and is projected back to the initial, uncoarsened, network, in the uncoarsening phase.
   Further details are given in Section~\ref{bac} and specificities related to these
-  phases when applied to visual analytics are in Sections~\ref{net} and~\ref{nav}.
+  phases when applied to interactive data visualization are in Sections~\ref{net} and~\ref{nav}.
   }\label{mlf}
@@
 i.e. $\forall\; l_1, l_2 \in M, u \in l_1 \Rightarrow u \notin l_2$.
-The heavy-edge mathing, for example, is a matching that fits this canonical description
+The heavy-edge matching, for example, is a matching that fits this canonical description
 in attempting to maximize total matching weight, but the matching may not satisfy such restrictions.
@@
 and the links incident to such nodes are joined into superlinks among the supervertices,
-also with total weight equal to the sum of the weight of the links merged.
-The solution finding is usually reasoned about in terms of the computational cost, being much
+also with total weight equal to the sum of the weights of the links merged.
+The solution finding phase is usually reasoned about in terms of the computational cost, being much
 alleviated in the reduced network.
@@
 \subsection{Network visualization assisted by multilevel strategies}\label{net}
-In applying the multilevel framework to visualization,
-the target algorithm is the visual mapping of the network,
+In applying the multilevel framework to visualization, adaptations are suitable to
+the achievement of the visualization intended.
+Such adaptations for our interactive visualization solution is as follows.
+The target algorithm is the visual mapping of the network,
 i.e. the solution finding phase is the visualization of the network by means of
 the reduced representation.
-The uncoarsening is performed only on demand, through user requests
-to visualize network regions with more detail, thus avoiding unecessary complexity and,
+The uncoarsening is performed only on demand through user requests
+to visualize network supernodes in more detail,
+thus avoiding unecessary complexity and,
 most importantly, avoiding to overload the visualization with information beyond the
@@
 The interactivity is crucial for a number of reasons:
-the definition of the uncoarsening phase desired;
-for the navigation of the network,
+the specification of the coarsening algorithm;
+the definition of the uncoarsening desired;
+to enable the navigation of the network,
 including the access to metadata and changes to the visualization achieved;
-for uncoarsening supernodes;
 and for tuning the achieved visual mapping status, such as by resizing
@@
   \quad maximum number of levels: $L \in [0,n] \subset \mathbb{Z}$\\
-  \quad reduction factor: $rf \in (0,0.5] \subset \mathbb{R}$.\\
-  \quad layers to be coarsened: $layers \in {1,2}$\\
+  \quad reduction factor for each layer: $rf_1, rf_2 \in (0,0.5] \subset \mathbb{R}$.\\
+  \quad layers to be coarsened: $layers \in \{1,2\}$\\
   \quad user command given through the visual interface: $C$
@@
     \While{$l\leq L$}{
-      $M \leftarrow$ \Matching{$G_l$, $i$, $rf$}\;
+      $M \leftarrow$ \Matching{$G_l$, $i$, $rf_1$, $rf_2$}\;
       $G_{l+1} \leftarrow$ \Contracting{$G_l$, $M$}\;
@@
   $V \leftarrow$ \map{$G_L$}\;
-  \While{1}{
+  \While{$C$ is not '$exit$'}{
     $C\leftarrow$ user command\;
@@
   The routines {\bf Matching$_b$} and {\bf Contracting$_b$} are any of the matching or contracting
-  algorithms suited for bipartite networks, such as described in~\cite{alan2} which are only ones
+  algorithms suited for bipartite networks, such as described in~\cite{alan2} which are the only ones
   currently available to the knowledge of the authors.
@@
   The visual mapping may then be transformed by the user by requesting uncoarsening of specific supernodes,
-  or by other commands not specific to the multilevel strategy, such as requesting metadata exposition, changes to the position of nodes, the color or transparency of nodes and links, zoom and pan,
+  or by other commands available to assist in the visualization,
+  such as requesting metadata exposition, changes to the position of nodes, the color or transparency of nodes and links, zoom and pan,
   or any other operation defined in Section~\ref{sof}.}\label{alg}
@@
 Accordingly, our idealized exploration pathway starts with an overview,
-achieved by the mapping the coarsest representation of the network through
+achieved by the visual mapping of the coarsest representation of the network through
 standard network layout algorithms for node-link diagrams.
@@
   \item reposition nodes, exposing linking patterns.
-  \item request metadata, such as number of children, the parent in a subsequent level, or the number of neighbors.
-  \item request visual mapping of network features, such as node size related to number of neighbors or the number of child nodes it represents.
+  \item request metadata, e.g. for a node, one may request to know the number of predecessors,
+    the successor in a subsequent level, or the number of neighbors (i.e. the degree).
+  \item request the visual mapping of network features, such as node size related to number of neighbors or the number of predecessors the node represents.
   \item request the uncoarsening of specific supernodes or any arbitrary group of supernodes.
@@
 network size, open supervertices, levels exposed, and topological features such as community structures.
+These operations are detailed in Section~\ref{sof}.
 The navigations pathway, already as achieved within the software we made available, is illustrated in Figure~\ref{fnav}.
@@
   and auxiliary tools.
+  Compliant with the Shneiderman's mantra, first and overview of the bipartite network is provided.
+  Then the user may select an specific portion of the network to visualize in further detail,
+  apply filters and visual mappings of topological features, and request details.
+  These operations may be carried on in any order and any number of times the analyst understand convenient.
   }\label{fnav}
@@
 \noindent 
-\section{Software implementation: mlvis}\label{sof}
-%%%
-% description of the functionalities/tools implemented,
-% of the technologies and libraries used
+\section{Software implementation}\label{sof}
 % optimization of the computational capacity for large networks by
@@
 and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
-The final result is exemplified in Figure~\ref{fpage} and next subsections
-describe its functionalities and the technologies used.
+The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization),
+is exemplified in Figures~\ref{fpage0} and~\ref{fpage}.
+% TTM
+Next subsections describe its functionalities and the technologies used.
+
+\begin{figure}[!h]\centering
+ \includegraphics[width=\textwidth]{fpage0}
+  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
+  and auxiliary tools.
+  }\label{fpage0}
+\end{figure}
 
@@
 
-\subsection{The use of mlvis}
+\subsection{The use of MlBiNetVis}
 % network upload, format.
@@
 % tools available and their operation
+In using MlBiNetVis,
+the analyst first selects a network of interest to be explored.
+One may also upload a network and then select it on the same drop-down menu.
+The user may then wish to select one of the available network layout
+algorithms and tune the coarsening to be performed, but default settings are reasonable
+for the newcommer.
+Upon request given by hiting the ``render network'' button,
+the network is mapped to the screen.
+Subsequent usage relies on manipulation of the visual mapping presented on the
+canvas by means of actions
+performed through managing the tools and related mouse operations.
+These steps are illustrated in Figure~\ref{initial}.
 
-\subsection{Technologies used within mlvis}
+% expose about text area, show links checkbox, layout button, and other elements. TTM
+
+There are multiple levels of representation entailed by the multilevel strategy.
+This makes pertinent that one and only of the available levels is explicitly selected
+so that the transformations are performed on that level.
+Thus, there is always a selected level in which the tools perform actions, such
+as link coloring or opening supernodes to expose their predecessor nodes.
+As can be infered by the reader,
+the number of nodes and links in each level which are visible are not fixed thoughout the
+navigation.
+The selection of the level to which the transformations should apply, and the number
+of visible and total nodes and links in each level, are exposed in the table to the right
+of the canvas, as in Figure~\ref{secondPhase}.
+In the same table, the user may change the color and shape of the nodes in each layer of
+a level, the color of the links of a level, and toggle show/hide of the links is a level,
+by left/right clicks on the corresponding colored cell.
+These operations are illustrated in Figure~\ref{using}.
+
+The toolbar holds buttons for further actions to be performed and are divided in groups.
+The first group is for changing how the nodes are visualized: size, proportionality to
+the number of neighbors or predecessors, reset proportionality, transparency, and rotation.
+The second group of buttons is for the links: thickness, transparency, proportionality to weight
+using thickness or transparency, and reset proportionality.
+
+The third group of buttons is more specific to the needs of the MlBiNetViz, and
+are selected to make mouse operations entail special processes and transformations of the visualization.
+The first button is selected for the interface to return core information from a clicked node:
+id, level, number of predecessors of previous level, id of successor (if node is not on the last level),
+number of neighbors (degree), sum of weight of links (strength).
+Other information the analysts report necessity of are easily implemented, but currently only these informations
+are returned to avoid unecessarily flooding the text area.
+The second tool button is seleted to move nodes: specific nodes are individually moved if clicked and dragged;
+a rectangular region is made available for moving all nodes therein if an empty region of the canvas
+is clicked and the mouse cursor dragged before releasing the button.
+The third tool button is selected for opening supernodes and exposing their predecessor nodes and links among them:
+a region is defined by the user by clicking on an empty space on the canvas and gragging the mouse before releasing,
+the supernodes therein are joined into a rectangular region that contains all of them,
+and their predecessor nodes (which are on a finer level of detail) are drawn inside such rectangular region.
+Open supernodes may be joined using the same tool, but it was found convenient to make available a
+fourth button to assist in this task: if this tool is selected, one may join two open supernodes by
+clicking on them in sequence.
+The last tool button in this group is selected to perform three operations on open supernodes:
+if an open supernode is clicked on and dragged, the rectangular shape is resized accordingly;
+if there is no drag, i.e. the super node is clicked on and released, the links from it to the other
+supernodes are attached to the next counterclockwise corner of its rectagular region.
+When the open supernode is resized, the layout of its child nodes is optimized for the drawn region.
+This is most useful when used in combination with the drag tool:
+on the finer level of detail, the nodes are repositioned into a more compact disposition.
+Then upon a (potentially minimal) supernode
+risize, the child nodes are made to fit the whole rectangular area.
+The last set of tools acts on the whole canvas:
+changing the color of the background, zoom, left-right/up-down pan,
+rotation of the entire network,
+and toggling between current zoom and pan status and the initial setting.
+The toolbar is contextualized in Figure~\ref{using} and their use is illustrated in Figure~\ref{tools}.
+
+There are many subleties on the orchestrated use of these tools.
+As a facilitated means to convey the navigation possibilities and
+procedures available, the user is invited to watch a demo video in~\cite{yvideo},
+and use the MlBiNetVis interface through the instance we made availabe~\cite{mbnvpage}.
+The more seasoned software developer may browse the code, upload a local instance
+of the software, and make changes to its source code~\cite{mbnvcode}.
+
+\subsection{Software technologies employed within mlvis}
 % coarsening algorithms
@@
 % Nuxt/Vue.js for the frontend with heavy use of PIXI.js
+Most important parts of MlBiNetViz are written in a combination of JavaScript and Python:
+Vue.js (set up by Nuxt.js) is used to assist in the client (frontend),
+the server (backend) is a Flask Python server, used to perform
+specialized or heavy calculations.
+There is a secondery server, a FeatherJS, used to facilitate contact with the database
+and real-time multi-user interaction.
+The data is stored in a MongoDB database and ordinarily in the filesystem,
+while the multi-user interaction was deactivated for now to avoid unecessary complexity.
+The coarsening algorithms for bipartite networks are available in~\cite{bialgs}
+and are used by the Flask server.
+The fast WebGL 2D rendering on the canvas is performed using Pixi.js.
 
-\section{Results and discussion}\label{res}
+The rasterization of images is usually performed though triangles.
+Accordingly, in order to comply with the goal of enabling the visualization of large bipartite networks,
+the rendered network on the canvas contains only triangles (one for each of the nodes) and straight lines (for the links).
+The user may wish to further alleviate the computational cost by not showing the links.
+To emphasize the bipartite structure of the network, and for making the visualization more aesthetically
+compelling, the user may change the shape of the nodes in one (or both) of the layers to hexagons instead of triangles.
+Using these features, and within the technologies described, we have visualized networks
+with tens of thousands of nodes without perceiving any lag in the interactive navigation and transformations,
+with links shown,
+even using ordinary machines (e.g. with 8GB RAM DDR3, an i7 processor of the first generations, and a 1GB GPU).
+
+This is a very brief account of the styles and software resources used to implement
+MlBiNetVis. The interested reader is welcome to browse the software code and documentation
+and contact the authors~\cite{mbvncode}.
+
+% \section{Results and discussion}\label{res}
 %%%
@@
 % extension to multipartite (or heterogeneous) networks.
+We have described and made available the first visualization interface for bipartite networks
+assited by multilevel strategies.
+The resulting conceptual organization is very consistent with the Shneiderman's mantra,
+as exemplified the by the implemented software.
+The interface developed is efficiently managed through simple interactive operations,
+and very efficient in visualization large networks due to the use of multilevel strategies,
+the software technologies employed, and the computationally inexpensive design decisions regarding
+the rendering of the node-link diagrams.
+In summary, this work is a proof-of-concept in that it establishes as feasible the use
+of multilevel strategies for the visualization of large bipartite networks.
+
+In further steps, the navigation may be enhanced by the acute use of in-development coarsening algorithms.
+The visualization pathway is being envisioned as capable to assist the development of novel
+multilevel strategies, which should be verified.
+Also, there are multilevel strategies that involve the selection of nodes as pivors
+to guide the coarsening procedure, a process to which there is no visual interface,
+and to which MlBiNetVis should be adapted.
+Also, the interface is envioned to assist the analysis of coarsening results. As in the current version of MlBiNetVis,
+the analyst is able to scrutinize linking patterns and overall network structure though the node-link diagram,
+and requesting information, but this should be enhanced by the display of histograms and global measures of each level.
+On the analysis of the networked structures, the navigation involving metadata, such as from genes and proteines, or as
+documents and authors, should result in a visual analytics interface for interactive knowledge discovery.
+Finally, bipartite networks are naturally extended to heterogeneous (or multipartite) networks,
+and thus all the content of this article, and should require further research and development for the achievement
+of concrete derivatives.
 

### 2019-05-28 paper.tex
feita primeira figura
terminada primeira versao do artigo

@@
 These issues may be aggravated in the case of bipartite networks,
-due to their sparsity and topological complexities~\cite{alan1}.
+due to their sparsity and topological complexities.
 Bipartite networks are comprised of two partitions of nodes,
@@
 from the matched nodes.
-There are several coarsening algorithms reported in the scientific literature~\cite{}, a few of them
-developed for bipartite networks~\cite{}.
-A detailed exposition of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
+There are several coarsening algorithms reported in the scientific literature~\cite{co1,co2}, a few of them
+developed for bipartite networks~\cite{alan2}.
+A detailed exposition of these algorithms is found on the bibliography and is beyond the scope of this article.
 Most importantly, we are interested in coarsening suited for bipartite networks,
@@
 
-Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{}.
+Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{visAn}.
 Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
@@
 which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships,
-and providing a manipulable medium for the exploration of the information of interest~\cite{}.
+and providing a manipulable medium for the exploration of the information of interest.
 For this present work to be better characterized as a visual analytics contribution, the methods described in this section and software described in Section~\ref{sof} should further enable the discovery of information, such as metadata of the networks (e.g.
@@
 \begin{figure}[!h]\centering
- \includegraphics[width=0.6\textwidth]{fnav}
-  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
+ \includegraphics[width=0.8\textwidth]{fnav___}
+  \caption{The navigation pathway of bipartite networks obtained through a multilevel strategy
   and auxiliary tools.
@@
 and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
-The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization),
-is exemplified in Figures~\ref{fpage0} and~\ref{fpage}.
-% TTM
+The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization)
+and is exemplified in Figures~\ref{fpage0},~\ref{initial} and~\ref{secondPhase}.
 Next subsections describe its functionalities and the technologies used.
@@
 \begin{figure}[!h]\centering
- \includegraphics[width=\textwidth]{fpage0}
-  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
-  and auxiliary tools.
+ \includegraphics[width=\textwidth]{overall____}
+  \caption{The MlBiNetVis interface. (1) is the dropdown menu for choosing the network to analyse or uploading new ones.
+  The gray box in (2) contains widgets to select and parametrize the coarsening algorithm.
+  (3) is the render button, which maps the network to the screen.
+  In (4) are auxiliary widgets for selecting the  layout, show/hide of links, and for the interface to output information requested by the user or useful to assist navigation.
+  In (5) is an interactive table which holds information about the levels and layers of the bipartite network and may be used to set the visualization on the canvas.
+  In (6) is the toolbar that enables the navigation of the network and fine-tuning the visualization.
+  In (7) is the canvas in which the network is drawn through node-link diagrams.
+  Section~\ref{use} describes the use of these elements in further detail.
   }\label{fpage0}
@@
 
-\begin{figure}[!h]\centering
- \includegraphics[width=\textwidth]{fpage}
-  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
-  and auxiliary tools.
-  }\label{fpage}
-\end{figure}
-
-\subsection{The use of MlBiNetVis}
+\subsection{The use of MlBiNetVis}\label{use}
 % network upload, format.
@@
 These steps are illustrated in Figure~\ref{initial}.
-
-% expose about text area, show links checkbox, layout button, and other elements. TTM
+\begin{figure}[!h]\centering
+ \includegraphics[width=\textwidth]{initial_}
+  \caption{The elements in the initial step of the MlBiNetVis usage.
+  The button at the top is for network upload and selection.
+  In the gray box, the user chooses and sets the coarsening algorithm as detailed in Appendix~\ref{ap}.
+  The blue button at the bottom is a drop-down menu where the user picks the node-link
+  layout algorithm.
+  The green button is for rendering the network to the canvas and initializing further widgets.
+  By hitting the green button, all the elements in this image is made disbled, except for the layout
+  button.
+  Further information is given in Section~\ref{use}.
+  }\label{initial}
+\end{figure}
 
@@
 by left/right clicks on the corresponding colored cell.
-These operations are illustrated in Figure~\ref{using}.
+
+\begin{figure}[!h]\centering
+ \includegraphics[width=\textwidth]{secondPhase}
+  \caption{The elements in second step of the MlBiNetVis usage, dedicated to navigation on the network and transformation of the visualization.
+  The user may change the layout used when supernodes are opened and the predecessors are then shown on the screen.
+  Links may be shown or hided for all the levels by using the checkbutton.
+  The text area assists in the output of information requested or useful for navigation.
+  The canvas holds the node-link representations of the levels, which are transformed using the tools in the toolbar and the table in the right-side.
+  Such table also makes evident key information about the status of the visualization.
+  Further information is given in Section~\ref{use}.
+  }\label{secondPhase}
+\end{figure}
 
@@
 and toggling between current zoom and pan status and the initial setting.
-The toolbar is contextualized in Figure~\ref{using} and their use is illustrated in Figure~\ref{tools}.
+The toolbar is contextualized in Figure~\ref{secondPhase}.
 
@@
 MlBiNetVis. The interested reader is welcome to browse the software code and documentation
-and contact the authors~\cite{mbvncode}.
+and contact the authors~\cite{mbnvcode}.
 
@@
 
+\appendix
+
+\section{Choice and fine-tuning of the coarsening algorithm}\label{ap}
+The elements in Figure~\ref{initial}.
+
 \bibliographystyle{splncs04}

### 2019-05-28 paper.tex
spell check

@@
 \subsection{Multilevel coarsening of bipartite networks}\label{bac}
-Before~\cite{alan2}, multilevel approaches were not directly aplicable to bipartite networks.
+Before~\cite{alan2}, multilevel approaches were not directly applicable to bipartite networks.
 Such work introduced novel and efficient matching and coarsening algorithms,
@@
 in attempting to maximize total matching weight, but the matching may not satisfy such restrictions.
-Selecting clicks and other larger node sets to be merged into supernodes are possibilites
+Selecting clicks and other larger node sets to be merged into supernodes are possibilities
 under development.
@@
 to visualize network supernodes in more detail,
-thus avoiding unecessary complexity and,
+thus avoiding unnecessary complexity and,
 most importantly, avoiding to overload the visualization with information beyond the
@@
 for use through simple mouse-driven actions
-and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
+and requiring the installation of no software but a web browser such as Firefox or Google Chrome.
 The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization)
@@
 algorithms and tune the coarsening to be performed, but default settings are reasonable
-for the newcommer.
-Upon request given by hiting the ``render network'' button,
+for the newcomer.
+Upon request given by hitting the ``render network'' button,
 the network is mapped to the screen.
@@
   The green button is for rendering the network to the canvas and initializing further widgets.
-  By hitting the green button, all the elements in this image is made disbled, except for the layout
+  By hitting the green button, all the elements in this image is made disabled, except for the layout
   button.
@@
 as link coloring or opening supernodes to expose their predecessor nodes.
-As can be infered by the reader,
-the number of nodes and links in each level which are visible are not fixed thoughout the
+As can be inferred by the reader,
+the number of nodes and links in each level which are visible are not fixed throughout the
 navigation.
@@
   The user may change the layout used when supernodes are opened and the predecessors are then shown on the screen.
-  Links may be shown or hided for all the levels by using the checkbutton.
+  Links may be shown or hided for all the levels by using the checkbox.
   The text area assists in the output of information requested or useful for navigation.
@@
 Other information the analysts report necessity of are easily implemented, but currently only these informations
-are returned to avoid unecessarily flooding the text area.
-The second tool button is seleted to move nodes: specific nodes are individually moved if clicked and dragged;
+are returned to avoid unnecessarily flooding the text area.
+The second tool button is selected to move nodes: specific nodes are individually moved if clicked and dragged;
 a rectangular region is made available for moving all nodes therein if an empty region of the canvas
@@
 The third tool button is selected for opening supernodes and exposing their predecessor nodes and links among them:
-a region is defined by the user by clicking on an empty space on the canvas and gragging the mouse before releasing,
+a region is defined by the user by clicking on an empty space on the canvas and dragging the mouse before releasing,
 the supernodes therein are joined into a rectangular region that contains all of them,
@@
 if there is no drag, i.e. the super node is clicked on and released, the links from it to the other
-supernodes are attached to the next counterclockwise corner of its rectagular region.
+supernodes are attached to the next counterclockwise corner of its rectangular region.
 When the open supernode is resized, the layout of its child nodes is optimized for the drawn region.
@@
 Then upon a (potentially minimal) supernode
-risize, the child nodes are made to fit the whole rectangular area.
+resize, the child nodes are made to fit the whole rectangular area.
 The last set of tools acts on the whole canvas:
@@
 
-There are many subleties on the orchestrated use of these tools.
+There are many subtleties on the orchestrated use of these tools.
 As a facilitated means to convey the navigation possibilities and
 procedures available, the user is invited to watch a demo video in~\cite{yvideo},
-and use the MlBiNetVis interface through the instance we made availabe~\cite{mbnvpage}.
+and use the MlBiNetVis interface through the instance we made available~\cite{mbnvpage}.
 The more seasoned software developer may browse the code, upload a local instance
@@
 specialized or heavy calculations.
-There is a secondery server, a FeatherJS, used to facilitate contact with the database
+There is a secondary server, a FeatherJS, used to facilitate contact with the database
 and real-time multi-user interaction.
-The data is stored in a MongoDB database and ordinarily in the filesystem,
-while the multi-user interaction was deactivated for now to avoid unecessary complexity.
+The data is stored in a MongoDB database and ordinarily in the file system,
+while the multi-user interaction was deactivated for now to avoid unnecessary complexity.
 The coarsening algorithms for bipartite networks are available in~\cite{bialgs}
@@
 We have described and made available the first visualization interface for bipartite networks
-assited by multilevel strategies.
+assisted by multilevel strategies.
 The resulting conceptual organization is very consistent with the Shneiderman's mantra,
@@
 multilevel strategies, which should be verified.
-Also, there are multilevel strategies that involve the selection of nodes as pivors
+Also, there are multilevel strategies that involve the selection of nodes as pivots
 to guide the coarsening procedure, a process to which there is no visual interface,
 and to which MlBiNetVis should be adapted.
-Also, the interface is envioned to assist the analysis of coarsening results. As in the current version of MlBiNetVis,
+Also, the interface is envisioned to assist the analysis of coarsening results. As in the current version of MlBiNetVis,
 the analyst is able to scrutinize linking patterns and overall network structure though the node-link diagram,
 and requesting information, but this should be enhanced by the display of histograms and global measures of each level.
-On the analysis of the networked structures, the navigation involving metadata, such as from genes and proteines, or as
+On the analysis of the networked structures, the navigation involving metadata, such as from genes and proteins, or as
 documents and authors, should result in a visual analytics interface for interactive knowledge discovery.

### 2019-05-28 paper.tex
arrumacoes minimas

@@
 \section{Method description}\label{des}
-\subsection{Fundamental concepts}\label{bac}
+\subsection{Fundamental concepts}
 A bipartite network $G=(V,E)$ consists in a set $V$ of nodes which is partitioned in two subsets
@@
 
-\subsection{Multilevel coarsening of bipartite networks}\label{bac}
+\subsection{Multilevel coarsening of bipartite networks}
 Before~\cite{alan2}, multilevel approaches were not directly applicable to bipartite networks.
@@
 
-\begin{figure}[!h]\centering
+\begin{figure}\centering
  \includegraphics[width=\textwidth]{mlf}
@@
   and is projected back to the initial, uncoarsened, network, in the uncoarsening phase.
-  Further details are given in Section~\ref{bac} and specificities related to these
+  Further details are given in Section~\ref{des} and specificities related to these
   phases when applied to interactive data visualization are in Sections~\ref{net} and~\ref{nav}.
@@
 
-\begin{figure}[!h]\centering
+\begin{figure}\centering
  \includegraphics[width=0.8\textwidth]{fnav___}
@@
 
-\begin{figure}[!h]\centering
+\begin{figure}\centering
  \includegraphics[width=\textwidth]{overall____}
@@
 These steps are illustrated in Figure~\ref{initial}.
-\begin{figure}[!h]\centering
+\begin{figure}\centering
  \includegraphics[width=\textwidth]{initial_}
@@
 
-\begin{figure}[!h]\centering
+\begin{figure}\centering
  \includegraphics[width=\textwidth]{secondPhase}
@@
 \bibliography{biblio}
-%
-% \begin{thebibliography}{8}
-% \bibitem{ref_article1}
-% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
-% 
-% \bibitem{ref_lncs1}
-% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-% Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-% 
-% \bibitem{ref_book1}
-% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-% Location (1999)
-% 
-% \bibitem{ref_proc1}
-% Author, A.-B.: Contribution title. In: 9th International Proceedings
-% on Proceedings, pp. 1--2. Publisher, Location (2010)
-% 
-% \bibitem{ref_url1}
-% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-% Oct 2017
-% \end{thebibliography}
-\end{document}

### 2019-06-12 paper.tex
revisado artigo

--- 2019-05-28 paper.tex
@@
 \usepackage{subfig}
+\usepackage{placeins}
 % Used for displaying a sample figure. If possible, figure files should
@@
 %
-\title{Interactive Visualization of Large Bipartite Networks Assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
+\title{Interactive Visualization of Large Bipartite Networks Assisted by Multilevel Strategies}
+%\thanks{Supported by FAPESP, project 2017/05838-3}}
 %
@@
 \author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
-Alan Valejo \and
-Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
+Alan Valejo\orcidID{0000-0002-9046-9499} \and Alneu A. Lopes\orcidID{0000-0003-3112-4746}
+Maria Cristina F. Oliveira\orcidID{0000-0002-4729-5104}}
 %
@@
 \email{alanvalejo@gmail.com},
-\email{cristina@icmc.usp.br}\\
-\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
+\email{alneu@icmc.usp.br}
+\email{cristina@icmc.usp.br}}
+%\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
 %
@@
 % 150--250 words.
-  It is a well established fact that bipartite, or two-layer, networks are pervasive
-  in modeling real-world phenomena and that they play fundamental roles in
+  Bipartite, or two-layer, networks are pervasive
+  in modeling real-world phenomena and play a fundamental role in
   graph theory.
-  Multilevel strategies have been developed for optimization tasks,
-  and for the visualization of simple (``unipartite'') networks,
-  but their employment for visualizing bipartite networks were not found by the authors.
-  In this work, we present advances in the use of multilevel strategies for the
-  visualization of bipartite networks,
-  allowing for interactive and intuitive navigation of such structures and visual mappings of large datasets.
-  More specifically, we developed a web-based visualization interface in which
-  a parametrizable simplification of
-  bipartite networks are obtained through the application of coarsening algorithms.
-  The resulting networks are then presented to the user,
-  providing a genuine route for the ``overview first - focus on demand''
-  process on the analysis of the underlying data, in which the analyst
-  selects supervertices or whole network sectors for more detailed observation,
-  i.e. performs requests for the interface to display 
-  specific structures in less simplified settings.
-  The application allows not only the visualization of large networks,
-  but also for an on-demand navigation of the multilevel structure,
-  and is useful for the further development multilevel strategies themselves
-  e.g. by the specification of vertices to guide the coarsening processes and
-  the examination of the resulting multilevel hierarchy.
+  Multilevel methods, introduced for solving optimization problems on networks, have been applied to the problem of drawing simple (``unipartite'') networks, but not for interactive visualization of bipartite networks.
+  In this work, we present a proof-of-concept implementation on the use of the multilevel method for this purpose.
+We developed a web-based visualization interface in which multilevel coarsening algorithms are applied to obtain a hierarchy of gradually simplified representations of an input  bipartite network.
+  Node-link representations of the resulting network hierarchy can be presented to users following a genuine route of the ``overview first, zoom and filter, details on demand'' visual information seeking mantra, as analysts may depart from a coarser representation and select super-nodes or sub-graphs (network partitions) to be expanded and shown at greater detail.
+    Such a solution allows for interactive and intuitive navigation of large-scale network structures and can provide the basis for more elaborate visual mappings of network data sets.
 
-\keywords{Network visualization \and Bipartite networks \and Multilevel strategies \and Big data \and Complex networks \and Data visualization.}
+  %The application allows not only the visualization of large networks, but also for an on-demand navigation of the multilevel structure,  and is useful for the further development multilevel strategies themselves e.g.
+by the specification of vertices to guide the coarsening processes and  the examination of the resulting multilevel hierarchy.
+
+\keywords{Network visualization \and Bipartite networks \and Multilevel strategies \and Big data \and Complex networks \and Data visualization}
 \end{abstract}
@@
 \section{Introduction}
-The visualization of large-scale networks poses challenges both in terms of computational costs
-and of effective presentation of the information to the user~\cite{tang,staudt}.
+The visualization of large-scale networks poses challenges in terms of both the computational cost and the effective presentation of information~\cite{tang,staudt}.
 These issues may be aggravated in the case of bipartite networks,
-due to their sparsity and topological complexities.
-Bipartite networks are comprised of two partitions of nodes,
-called ``layers'',
-and links are not incident between nodes in the same partition.
-Such network type arises very often and naturally from the representation
-of relations among two kinds objects,
-e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
-Furthermore, real-world networks are often bipartite, and most unipartite networks
-are projections of bipartite networks or exhibit bipartite properties~\cite{guillaume0,guillaume}.
-In order to assist the visualization and navigation of large networks, one possibility is the use of
-multilevel strategies, which consist on the employment of incremental coarsening of the original
-network to obtain a sequence of simplified representations, i.e. with fewer nodes and links.
-Multilevel strategies are most traditionally used for executing complex algorithms
-on large-scale networks by the application of the algorithm on a smaller
-version of the network~\cite{alan2,ml2}.
-The employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
-networks has been reported, but their exploitation for visualizing bipartite networks
-was not found by the authors, as described in the next subsection.
-Accordingly, we present a system for the visualization of bipartite networks using
-multilevel strategies developed for bipartite networks.
-The system consists on presenting a simplified version of the network to the user, which then
-requests for supervertices (or collections of them) to be uncoarsened and visualized in more detail.
-Thus, the resulting interface allows for the visualization of large structures;
-enables an interactive navigation of the structure e.g. by uncoarsening supernodes, zooming and requesting information about the nodes;
-and may assist not only in the analysis of bipartite networks, but also on the analysis and development of multilevel algorithms in bipartite networks.
-Most importantly, the overall method, and the implemented interface, is a proof-of-concept, it demonstrates the feasibility of the use of multilevel strategies for the efficient visualization of large bipartite networks.
+due to their sparsity and topological complexity.
+In bipartite networks the set of nodes is split into two partitions called ``layers''
+and links are not allowed between nodes in the same partition.
+Such networks arise often and naturally from the representation
+of relations between entities of two kinds,
+e.g. documents and terms, papers and authors~\cite{doc,sci,movie}, or patients and genes~\cite{gene}.
+Furthermore, real-world networks are often bipartite and most unipartite networks are actually projections of bipartite networks or exhibit bipartite properties~\cite{guillaume0,guillaume}.
+%Multilevel strategies can be employed to assist the visualization and navigation of large networks, which consist on obtaining an incremental coarsening of the original network to yield a sequence of gradually simplified representations, i.e.
+with fewer nodes and links.
 
-This paper is organized as follows: in Section~\ref{rel} the related work is examined,
-while in~\ref{nom} are selected remarks about the vocabulary.
-The method is delineated in Section~\ref{des} and
-the software implementation is then described in Section~\ref{sof}.
-% Results and discussion are in~\ref{res}.
-Finally, Section~\ref{con} holds conclusions and further work envisioned.
+Multilevel strategies (see Section~\ref{des}) perform an incremental coarsening of an original
+network to yield a sequence of gradually simplified representations, i.e. with fewer nodes and links.
+They have been traditionally employed to enable executing expensive algorithms
+on large-scale networks\footnote{By large-scale we mean networks with up to a few hundreds of thousands vertices, in contrast to massive-scale networks with  millions to billions vertices.}: the rationale is to run the algorithm on a coarsened
+version that preserves the major properties of the input network to compute an initial solution, which is projected into the inverse hierarchy of coarsened networks  to yield a solution to the problem relative to the original network~\cite{alan2,ml2}.
+The method has been mostly applied to identify community structures in large networks~\cite{Pope2017}, but also to other problems, including computing node-link layouts of simple (i.e.
+``unipartite'') networks~\cite{u1,u2,u3,u4,u5,u6,u7,u8,u9,u10}.
+We are aware of a single (yet unpublished) contribution in the context of bipartite networks, in which a visual metaphor is used to assist users executing multilevel methods~\cite{cintra2019}.
+In general, solutions for node-link visualizations of large networks often rely on aggregation of nodes in clusters or communities to reduce visual clutter and create simpler representations amenable to user interaction~\cite{a1,a2,a3,a4,a5,a6}.
 
-\subsection{Related work}\label{rel}
+In this paper we present a system that employs the multilevel method to enable interactive visualization of large bipartite networks with node-link representations.
+Its underlying rationale is to present first a coarser instance of the network with which a user can interact to request for super-nodes (or groups of them) to be uncoarsened and visualized in more detail.
+The resulting interface enables interactive navigation on large networks, e.g.
+by gradual and on-demand uncoarsening of super-nodes, combined with functionalities for zooming and requesting complementary information on nodes and links.
+Besides assisting in the analysis of data represented as bipartite networks, the visualization may assist developers of multilevel algorithms, e.g., for inspecting and comparing results yielded by different algorithm and parameter choices.
+%Visual analytics is the field dedicated to the science of analytical reasoning assisted by interactive visual interfaces~\cite{visAn}.
+It is specially concerned with coupling interactive visual representations to support sense and decision making.
+The visual analytics techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships, and providing a manipulable medium for exploring and identifying the information of interest.
+%For this solution to be properly characterized as a visual analytics contribution, the methods and software described in Sections~\ref{des} and~\ref{sof} should further enable the discovery of information, such as metadata associated with  network nodes and links (e.g.
+authors, words or documents).
+%Accordingly, our methods and software may be understood as data visualization with explicit visual analytics elements, such as interactivity, on-demand supply of information, and selective navigation.
+We consider the current implementation as a proof-of-concept to demonstrate the feasibility of using the multilevel method to support effective visualization of large bipartite networks using node-link views.
+%Most importantly, the overall method, and the implemented interface, is a proof-of-concept, it demonstrates the feasibility of the use of multilevel strategies for the efficient visualization of large bipartite networks.
+
+%This paper is organized as follows. 
+%in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
+%The method is delineated in Section~\ref{des} and the software implementation is described in Section~\ref{sof}.
+% Results and discussion are in~\ref{res}. Finally, Section~\ref{con} holds conclusions and further work envisioned.
+
+%\subsection{Related work}\label{rel}
 %%%
@@
 % tratar mais especificamente de estratégias multinivel?
-Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
-Also, the aggregation of clusters have been reported, and comprises an approach that resembles
-the coarsening procedure in creating simplified representations of the original network~\cite{a1,a2,a3,a4,a5,a6}.
-Even so, the authors are not aware of previous reports on the use of multilevel strategies
-for the visualization and navigation of bipartite networks.
-Furthermore, only~\cite{a1,a2,a3,a4,a6} report on interactive systems for network visualization,
-and these works do not tackle the use of multilevel strategies for visualization.
+%Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
+%Also, the aggregation of clusters has been reported, an approach that resembles a coarsening procedure in creating simplified representations of an input network~\cite{a1,a2,a3,a4,a5,a6}.
+%We are not aware of previous reports on the use of multilevel strategies for the visualization and navigation of bipartite networks.
+%Furthermore, few contributions~\cite{a1,a2,a3,a4,a6} report on interactive systems for network visualization, and these do not tackle the usage of multilevel strategies for this purpose.
 
-\subsection{Nomenclature remarks}\label{nom}
+%\subsection{Remarks on nomenclature}\label{nom}
 %%%
@@
 % multilevel strategies synonym for optimization algorithms?
-The main vocabulary issue that arises in the context of this article is between
-level and layer. Layers are the (two) node partitions of a bipartite network, while levels comprise the sequence
-of coarsened networks.
-Also, simple or homogeneous networks are opposed to heterogeneous networks, in which there
-are more then one type of node.
-Bipartite networks may be regarded as an elementary case of heterogeneous networks,
-but it has one additional restriction: only nodes of different types are connected.
-Another usual concept in this context is that of a multilayer network, in which there are layers,
-i.e. partitions of nodes, in addition to nodes and links.
-Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layer.
-Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
-A thorough exposition of the vocabulary is beyond the scope of this article, but
-some attention for the issue is helpful to assist searches and newcomers.
+%The main vocabulary issue that arises in the context of this article is between level and layer.
+Layers are the (two) node partitions of a bipartite network, while levels comprise the sequence of coarsened networks.
+%Also, simple or homogeneous networks are opposed to heterogeneous networks, in which there are more then one type of node.
+%Bipartite networks may be regarded as an elementary case of heterogeneous networks, but it has one additional restriction: only nodes of different types are connected.
+%Another usual concept in this context is that of a multilayer network, in which there are layers, i.e. partitions of nodes, in addition to nodes and links.
+%Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layer.
+%Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
+%A thorough exposition of the vocabulary is beyond the scope of this article, but some attention for the issue is helpful to assist searches and newcomers.
 
-\section{Method description}\label{des}
-\subsection{Fundamental concepts}\label{bac}
+\section{A brief description of the multilevel method}\label{des}
+
+In this section we address two topics.
+First we introduce some fundamental concepts and present an overview of the multilevel method in the context of bipartite networks.
+Then we outline how it can be instantiated to support interactive navigation  on such networks.
+
+%\subsection{Fundamental concepts}
+\subsection{Multilevel coarsening of bipartite networks}\label{smul}
 A bipartite network $G=(V,E)$ consists in a set $V$ of nodes which is partitioned in two subsets
 with no links between nodes in the same set, i.e. $\exists\; V_1, V_2:\; V_1\cup V_2 = V,\;V_1\cap V_2 = \emptyset$, and $E\subseteq V_1 \times V_2$.
-The variable names are borrowed from more traditional
-Graph (network) theory, with Vertices (nodes) and Edges (links).
+The variable names are borrowed from traditional
+graph (network) theory, with Vertices (nodes) and Edges (links).
 One may regard the network as $G=(V,E,\sigma,\omega)$, with
@@
 
-A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network
-and where $|V_i| \geq |V_{i+1}|$.
-The coarsening procedure requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
-from the matched nodes.
-There are several coarsening algorithms reported in the scientific literature~\cite{co1,co2}, a few of them
-developed for bipartite networks~\cite{alan2}.
-A detailed exposition of these algorithms is found on the bibliography and is beyond the scope of this article.
-Most importantly, we are interested in coarsening suited for bipartite networks,
-as it has yielded simplified networks which present essential topological features
-of the original networks,
-and such result is perceived in the visualization as hinted in~\cite{alan2} and further shown in
-Section~\ref{sof}.
+%The multilevel optimization method is a meta-heuristics employed to guide, modify and potentially fix a solution obtained from a target algorithm. 
+%In the context of network problems, the method creates a hierarchy of coarsened networks $G_l$, $l$ integer $ \in [0,L]$ where $G_0$ is the original network and $|V_i| \geq |V_{i+1}|$.
+It operates in three phases, as illustrated in Figure~\ref{mlf}: coarsening, solution finding and uncoarsening, where a solution found in the coarsest network instance $G_L$ is projected back to the original, uncoarsened, network $G_0$.
 
-Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{visAn}.
+%The coarsening procedure requires two algorithms, the \emph{matching}, which defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation from the matched nodes.
+%Several coarsening algorithms have been reported in the literature~\cite{co1,co2}, a detailed exposition is beyond the scope of this paper.
+
+%and a few have been recently introduced for bipartite networks~\cite{alan2}.
+
+%, it has been shown that existing coarsening algorithms for bipartite networks can yield simplified models that preserve the essential topological features of the original networks~\cite{alan2}, as discussed further in Section~\ref{sof}.
+
+%Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{visAn}.
 Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
-Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways,
-which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships,
-and providing a manipulable medium for the exploration of the information of interest.
-For this present work to be better characterized as a visual analytics contribution, the methods described in this section and software described in Section~\ref{sof} should further enable the discovery of information, such as metadata of the networks (e.g.
+Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships, and providing a manipulable medium for the exploration of the information of interest.
+%For this present work to be better characterized as a visual analytics contribution, the methods described in this section and software described in Section~\ref{sof} should further enable the discovery of information, such as metadata of the networks (e.g.
 authors, words or documents).
-Accordingly, our methods and software may be understood as data visualization with explicit visual analytics elements, such as interactivity, on-demand supply of information, and selective navigation.
+%Accordingly, our methods and software may be understood as data visualization with explicit visual analytics elements, such as interactivity, on-demand supply of information, and selective navigation.
 
-\subsection{Multilevel coarsening of bipartite networks}\label{bac}
-Before~\cite{alan2}, multilevel approaches were not directly aplicable to bipartite networks.
-Such work introduced novel and efficient matching and coarsening algorithms,
-and scrutinized their validity for solving optimization problems, dimensionality reduction,
-and in the preservation of essential topological features of the original network.
-We here delineate such procedures in the context of our network visualization contribution.
+%\subsection{Multilevel coarsening of bipartite networks}
+The multilevel optimization method is a meta-heuristics employed to guide, modify and potentially fix a solution obtained from a target algorithm. 
+It operates in three phases, as illustrated in Figure~\ref{mlf}, namely coarsening, solution finding and uncoarsening.
+Valejo et al.~\cite{alan2} introduced a general framework applicable to bipartite networks.
+They presented novel and efficient matching and coarsening algorithms and validated them in problems of community detection and dimensionality reduction.
+They also provided empirical evidence that the proposed coarsening algorithms preserve the essential topological features of an original input bipartite network.
+Here we delineate the stages of the multilevel method in the context of our bipartite network visualization contribution.
 
-The multilevel optimization is a metaheuristics used to guide,
-modify and potentially fix a solution obtained from a target algorithm.
-It is divided into three phases: coarsening, solution finding and uncoarsening,
-where the solution found in the coarsest network $G_L$ is projected back to the original,
-uncoarsened, network $G_0$.
-Figure~\ref{mlf} illustrates such process.
+%The multilevel optimization method is a metaheuristics employed to guide, modify and potentially fix a solution obtained from a target algorithm.
+%It is split into three phases, as illustrated in Figure~\ref{mlf}: coarsening, solution finding and uncoarsening, where a solution found in the coarsest network instance $G_L$ is projected back to the original, uncoarsened, network $G_0$.
 
-\begin{figure}[!h]\centering
+%Figure~\ref{mlf} illustrates such process.
+
+\begin{figure}\centering
  \includegraphics[width=\textwidth]{mlf}
-  \caption{Phases of the multilevel optimization.
+  \caption{Phases of the general multilevel optimization method.
   A sequence of incrementally reduced networks is obtained in the coarsening phase.
-  An initial solution found at the coarsest network, in the solution finding phase,
-  and is projected back to the initial, uncoarsened, network, in the uncoarsening phase.
-  Further details are given in Section~\ref{bac} and specificities related to these
-  phases when applied to interactive data visualization are in Sections~\ref{net} and~\ref{nav}.
+  An initial solution is found applying a target algorithm in the coarsest network instance in the solution finding phase.
+This solution is projected back to the initial, uncoarsened, network in the uncoarsening phase.
+  %Further details are given in Section~\ref{des} and specificities related to these phases when applied to interactive data visualization are in Sections~\ref{net} and~\ref{nav}.
   }\label{mlf}
@@
 
-The coarsening phase results in a sequence of networks $G_l$ from the initial network $G_0$,
-on multiple levels of details.
+The coarsening phase creates a hierarchy of coarsened networks $G_l$, $l$ integer $ \in [0,L]$ where $G_0$ is the original network and $|V_i| \geq |V_{i+1}|$, resulting in a hierarchical representation of $G_0$ at decreasing levels of detail.
 The process is carried out by two algorithms: \emph{matching}, which decides which nodes to merge,
-and \emph{contracting}, which achieves the reduced representation from the network and the matching.
-In general, pairs of nodes are selected to be merged into supernodes, and
-most often a matching $M$ consists in a set of non-adjacent links,
-i.e. $\forall\; l_1, l_2 \in M, u \in l_1 \Rightarrow u \notin l_2$.
-The heavy-edge matching, for example, is a matching that fits this canonical description
-in attempting to maximize total matching weight, but the matching may not satisfy such restrictions.
-Selecting clicks and other larger node sets to be merged into supernodes are possibilites
-under development.
+and \emph{contracting}, which computes a reduced representation from the input network and the matching.
+In general, pairs of nodes are selected to be merged into super-nodes, and
+most often a matching $M$ consists of a set of non-adjacent links,
+i.e. $\forall\; l_1, l_2 \in M, u \in l_1 \Rightarrow u \notin l_2$~\cite{co1,co2}.
+%The heavy-edge matching, for instance, fits this canonical description attempting to maximize total matching weight, but the matching may not satisfy such restrictions.
+%Matching strategies based on selecting cliques or other larger node sets to be merged into super-nodes are possibilities under development.
 %%%
 % citar alguem aqui?
-For bipartite network, we use the algorithms provided by~\cite{alan2} in which two restrictions
-are imposed to the matching:
+For bipartite networks, we rely on the matching algorithms introduced by Valejo et al.~\cite{alan2} in which the matching must satisfy two restrictions:
+
 \begin{itemize}
   \item A node may only match nodes on the same layer.
-  \item A node may only match nodes that are reachable by two successive links
-    (i.e. the closest possible nodes on the same layer).
+  \item A node may only match nodes that are reachable in two hops, i.e., via paths formed by two successive links    (i.e.  the closest possible nodes in the same layer).
 \end{itemize}
@@
 The matching is followed by the contraction of the network into a coarser form.
-Typically, the nodes matched are merged into a supernode
-with the weight equal to the sum of the weight of nodes it contains,
-and the links incident to such nodes are joined into superlinks among the supervertices,
-also with total weight equal to the sum of the weights of the links merged.
-The solution finding phase is usually reasoned about in terms of the computational cost, being much
-alleviated in the reduced network.
-The uncoarsening is usually performed through the complete hierarchy,
-from the coarsest to the original network, with successive refinement of the solutions to
-avoid local minima and improve solution quality.
-These two phases are substantially different when the multilevel strategy is applied to visualization,
-which motivated their separate exposition in the next subsection.
+Typically, the nodes matched are merged into a super-node weighted according to the number of of its composing nodes, and the links incident to such nodes are joined into super-links,
+weighted based on the accumulated weights of the links merged.
+The solution finding phase usually involves executing a computationally expensive algorithm, which becomes feasible in the coarsest network.
+In the uncoarsening phase the solution found in the coarsest network instance $G_L$ is projected back to the original, uncoarsened, network $G_0$.
+Uncoarsening is thus performed across the complete hierarchy, from $G_L$ to $G_0$ with successive refinement of the solutions to avoid local minima and improve solution quality.
+%These two phases are substantially different when the multilevel strategy is applied to visualization, which motivated their separate exposition in the next subsection.
 
-\subsection{Network visualization assisted by multilevel strategies}\label{net}
-In applying the multilevel framework to visualization, adaptations are suitable to
-the achievement of the visualization intended.
-Such adaptations for our interactive visualization solution is as follows.
-The target algorithm is the visual mapping of the network,
-i.e. the solution finding phase is the visualization of the network by means of
-the reduced representation.
-The uncoarsening is performed only on demand through user requests
-to visualize network supernodes in more detail,
-thus avoiding unecessary complexity and,
-most importantly, avoiding to overload the visualization with information beyond the
-cognitive convenience of the user and the computational power that the machine being used
-is able to provide for real-time interactivity.
+%\subsection{Network visualization assisted by multilevel strategies}\label{net}
+In applying the general multilevel framework to obtain an interactive visualization solution, some adaptations are in order.  
+%Such adaptations for our interactive visualization solution is as follows.
+In this scenario the goal is to obtain a visual mapping of the network, i.e.
+the solution finding phase corresponds to computing a layout of $G_L$ suitable for presentation.
+Uncoarsening is performed  on-demand, through specific user requests
+to expand selected network super-nodes for more detailed visualization, which is possible displaying them at the subsequent hierarchical level.
+The procedure is delineated in the BiNetVis algorithm (Algorithm~\ref{alg}) and further detailed in Sections~\ref{nav} and~\ref{sof}.
 
-The interactivity is crucial for a number of reasons:
-the specification of the coarsening algorithm;
-the definition of the uncoarsening desired;
-to enable the navigation of the network,
-including the access to metadata and changes to the visualization achieved;
-and for tuning the achieved visual mapping status, such as by resizing
-and moving nodes.
-The overall procedure is delineated in Algorithm~\ref{alg} and further
-detailed in Sections~\ref{nav} and~\ref{sof}.
+This solution avoids unnecessary complexity and, most importantly, avoids overloading the visualization with information beyond the cognitive convenience of the user and the computational power available to ensure real-time interactivity.
+Users can interact to specify the desired coarsening algorithm, select the node-link drawing algorithm, select the target super-nodes for uncoarsening, navigate the network, access data and metadata, tune the visual mapping options, for instance to modify colors and resize and move nodes.
+%Interactivity is crucial for a number of reasons: the specification of the coarsening algorithm; the definition of the uncoarsening desired; to enable the navigation of the network, including the access to metadata and changes to the visualization achieved; and for tuning the achieved visual mapping status, such as by resizing and moving nodes.
+%The procedure is delineated in Algorithm~\ref{alg} and further detailed in Sections~\ref{nav} and~\ref{sof}.
 
@@
 \begin{algorithm}
+  \qquad\qquad\qquad{\bf\emph{BiNetVis Algorithm}}\\
+  \vspace{0.3cm}
   \KwIn{\\
@@
   }
-  \caption{An algorithmic description of the multilevel strategy adapted for the visualization of bipartite networks.
-  The routines {\bf Matching$_b$} and {\bf Contracting$_b$} are any of the matching or contracting
-  algorithms suited for bipartite networks, such as described in~\cite{alan2} which are the only ones
-  currently available to the knowledge of the authors.
-  The {\bf map\_to\_screen} routine is performed through the use of a network layout algorithm and
-  then the rendering of the network to the screen.
-  The visual mapping may then be transformed by the user by requesting uncoarsening of specific supernodes,
-  or by other commands available to assist in the visualization,
-  such as requesting metadata exposition, changes to the position of nodes, the color or transparency of nodes and links, zoom and pan,
-  or any other operation defined in Section~\ref{sof}.}\label{alg}
+  \caption{
+An algorithmic description of the multilevel method adapted for the visualization of bipartite networks, named the {\bf \emph{BiNetVis Algorithm}}.
+  The routines {\bf Matching$_b$} and {\bf Contracting$_b$} refer to any matching or contracting
+  algorithms applicable to bipartite networks~\cite{alan2}.
+  The routine {\bf map\_to\_screen} may refer to any node-link layout algorithm followed by a rendering of the network to the screen.
+  The user may then modify the visual mapping, e.g., request to uncoarsen selected super-nodes or use any of the other interaction commands available,
+  such as request metadata, change node position, change node and link color or transparency, perform zoom and pan, etc., as described in Section~\ref{sof}.}\label{alg}
 \end{algorithm}
@@
 \subsection{The navigation pathway}\label{nav}
-The abundance of information within large networks makes pertinent the
-application of
-the ``visual information-seeking mantra'', also known as Shneiderman's mantra:
+The abundance of information within large-scale networks makes pertinent the
+application of the well-known ``visual information-seeking mantra''~\cite{shneiderman1996eyes}:
 \emph{overview first, zoom and filter, then details-on-demand}.
-This mantra comprises a number of visual design guidelines,
-such as the details-on-demand technique, and provides
-a very acknowledged framework for designing information visualization applications.
-Accordingly, our idealized exploration pathway starts with an overview,
-achieved by the visual mapping of the coarsest representation of the network through
-standard network layout algorithms for node-link diagrams.
-The user may then zoom into specific regions of the network, and then request
-details by a number of operations, most importantly:
+This mantra embeds a number of visual design guidelines,
+such as the details-on-demand interaction strategy, and provides
+a widely acknowledged framework for designing information visualization applications.
+The navigation pathway supported by the current implementation of the proposed solution is illustrated in Figure~\ref{fnav}.
+Accordingly, exploration starts with an overview yielded by the visual mapping of the coarsest network instance, obtained using standard layout algorithms for node-link diagrams.
+The user may then zoom into specific network regions and request
+details using several operations, most importantly:
 \begin{itemize}
-  \item reposition nodes, exposing linking patterns.
-  \item request metadata, e.g. for a node, one may request to know the number of predecessors,
-    the successor in a subsequent level, or the number of neighbors (i.e. the degree).
-  \item request the visual mapping of network features, such as node size related to number of neighbors or the number of predecessors the node represents.
-  \item request the uncoarsening of specific supernodes or any arbitrary group of supernodes.
+  \item Reposition nodes, expose linking patterns.
+  \item Request metadata, e.g. for a node one may ask for its degree, or how many predecessors it has in the previous level, or how many successors in the subsequent level.
+  \item Adjust the visual mapping of network features, such as map the size of glyphs representing nodes to node degree or to the number of node predecessors, or modify link transparency.
+  \item Request the uncoarsening (i.e., expansion to the upper level) of a selected super-node or group of super-nodes.
 \end{itemize}
 
-Other operations are convenient to make the visual mapping adequate for the diverse settings possible:
-network size, open supervertices, levels exposed, and topological features such as community structures.
-These operations are detailed in Section~\ref{sof}.
-The navigations pathway, already as achieved within the software we made available, is illustrated in Figure~\ref{fnav}.
+Other operations are convenient to adequate the visual mapping to the current user needs, e.g., setting the network size, expanding super-nodes, exposing levels and highlighting topological features.
+These operations are further detailed in Section~\ref{sof}.
 
-\begin{figure}[!h]\centering
- \includegraphics[width=0.8\textwidth]{fnav___}
-  \caption{The navigation pathway of bipartite networks obtained through a multilevel strategy
-  and auxiliary tools.
-  Compliant with the Shneiderman's mantra, first and overview of the bipartite network is provided.
-  Then the user may select an specific portion of the network to visualize in further detail,
-  apply filters and visual mappings of topological features, and request details.
-  These operations may be carried on in any order and any number of times the analyst understand convenient.
+\begin{figure}\centering
+ \includegraphics[width=0.7\textwidth]{fnav____}
+  \caption{The navigation pathway on bipartite networks made possible using the multilevel method coupled with auxiliary interaction tools.
+  Compliant with the ``visual information seeking mantra''~\cite{shneiderman1996eyes}, initially it displays a view of the bipartite network in its coarsest form.
+  The user may then zoom-in into regions and select specific nodes or all nodes within a region to inspect and visualize in further detail, apply filters and modify the visual mappings of topological elements, or request to see the corresponding nodes in further detail, drilling-down to the next network level.
+  These operations are user-driven and may be carried out in any sequence.
   }\label{fnav}
@@
 We implemented the framework described in Section~\ref{des} using scientific,
-database and web resources,
-and made the final result available to the user within a web page
-for use through simple mouse-driven actions
-and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
-The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization)
+database and web resources, making it available within a web page
+for use through simple mouse-driven actions and requiring no software installation beyond a web browser. % such as Firefox or Google Chrome.
+The software has been named BiNetVis (from Multilevel Bipartite Network Visualization)
 and is exemplified in Figures~\ref{fpage0},~\ref{initial} and~\ref{secondPhase}.
-Next subsections describe its functionalities and the technologies used.
+In the following we describe its functionalities and the underlying technologies.
 
-\begin{figure}[!h]\centering
- \includegraphics[width=\textwidth]{overall____}
-  \caption{The MlBiNetVis interface. (1) is the dropdown menu for choosing the network to analyse or uploading new ones.
-  The gray box in (2) contains widgets to select and parametrize the coarsening algorithm.
-  (3) is the render button, which maps the network to the screen.
-  In (4) are auxiliary widgets for selecting the  layout, show/hide of links, and for the interface to output information requested by the user or useful to assist navigation.
-  In (5) is an interactive table which holds information about the levels and layers of the bipartite network and may be used to set the visualization on the canvas.
-  In (6) is the toolbar that enables the navigation of the network and fine-tuning the visualization.
-  In (7) is the canvas in which the network is drawn through node-link diagrams.
-  Section~\ref{use} describes the use of these elements in further detail.
+\begin{figure}\centering
+ \includegraphics[width=\textwidth]{overall2_}
+  \caption{The BiNetVis interface: (1) drop-down menu for selecting or uploading a network for analysis;
+  (2) control widgets to select and parameterize the coarsening algorithm;  (3) the render button to display the network; (4) auxiliary widgets for selecting the network layout, show/hide links, and area to display user requested information on nodes or links; (5) interactive table holding information about the multilevel hierarchy levels and the bipartite network layers; can also be used to set the visualization on the canvas; (6) toolbar with controls for navigation of the network and fine-tuning the visualizationi, better inspected in Figure~\ref{toolbar} of Appendix C; (7) canvas with the node-link representation of the multilevel network.
+Notice that nodes shown belong to four distinct hierarchical levels: level 3 (dark red/green), level 2 (pink/cyan), level 3 (navy blue/yellow) and level 2 (red/bright green).
+The current interaction focus in on level 2, as indicated by the orange mark.
+ %Elements are described in further detail in Section~\ref{use}.
   }\label{fpage0}
@@
 
-\subsection{The use of MlBiNetVis}\label{use}
+\subsection{Using BiNetVis}\label{use}
 % network upload, format.
@@
 % tools available and their operation
-In using MlBiNetVis,
-the analyst first selects a network of interest to be explored.
-One may also upload a network and then select it on the same drop-down menu.
-The user may then wish to select one of the available network layout
-algorithms and tune the coarsening to be performed, but default settings are reasonable
-for the newcommer.
-Upon request given by hiting the ``render network'' button,
-the network is mapped to the screen.
-Subsequent usage relies on manipulation of the visual mapping presented on the
-canvas by means of actions
-performed through managing the tools and related mouse operations.
-These steps are illustrated in Figure~\ref{initial}.
-\begin{figure}[!h]\centering
+In using BiNetVis, the analyst first uses a drop-down menu to select a network of interest, s/he may also upload a new network using the same menu.
+The next step is to select one of the available network layout algorithms and possibly tune the coarsening algorithm, as illustrated in Figure~\ref{initial}, although default settings are reasonable for a newcomer.
+Once the ``render network'' button is hit the network node-link representation is mapped to the screen according to the selected layout algorithm.
+Subsequent usage relies on manipulation of the visual mapping controls presented on the
+canvas by means of user actions and related mouse operations.
+These steps are illustrated in Figure~\ref{fpage0}.
+
+\begin{figure}\centering
  \includegraphics[width=\textwidth]{initial_}
-  \caption{The elements in the initial step of the MlBiNetVis usage.
-  The button at the top is for network upload and selection.
-  In the gray box, the user chooses and sets the coarsening algorithm as detailed in Appendix~\ref{ap}.
-  The blue button at the bottom is a drop-down menu where the user picks the node-link
-  layout algorithm.
-  The green button is for rendering the network to the canvas and initializing further widgets.
-  By hitting the green button, all the elements in this image is made disbled, except for the layout
-  button.
-  Further information is given in Section~\ref{use}.
+  \caption{The initial step of the BiNetVis usage is to apply the multilevel coarsening to an input network.
+  The button at the top is for network uploading and selection.
+  In the gray box, the user chooses the coarsening algorithm and sets its parameters, as detailed in the Appendix A. % Appendix~\ref{ap}.
+  The blue button at the bottom is a drop-down menu where the user can select the node-link layout algorithm, where multiple alternatives are provided.
+  The green button renders the network to the canvas and initializes further control widgets.
+  By hitting the green button, except for the layout  button all the control elements above are disabled as the interface enters the visualization mode.
+  %Further information is given in Section~\ref{use}.
   }\label{initial}
@@
 
-There are multiple levels of representation entailed by the multilevel strategy.
-This makes pertinent that one and only of the available levels is explicitly selected
-so that the transformations are performed on that level.
-Thus, there is always a selected level in which the tools perform actions, such
-as link coloring or opening supernodes to expose their predecessor nodes.
-As can be infered by the reader,
-the number of nodes and links in each level which are visible are not fixed thoughout the
-navigation.
-The selection of the level to which the transformations should apply, and the number
-of visible and total nodes and links in each level, are exposed in the table to the right
-of the canvas, as in Figure~\ref{secondPhase}.
-In the same table, the user may change the color and shape of the nodes in each layer of
-a level, the color of the links of a level, and toggle show/hide of the links is a level,
-by left/right clicks on the corresponding colored cell.
+The multilevel method yields a hierarchical representation of the network at decreasing levels of detail (as depicted in Figure~\ref{mlf}), where the number of hierarchy levels is an input parameter to the coarsening algorithm.
+The proposed visualization initially shows a node-link representation of the the coarsest network $G_L$ (the upper level instance) showing it at the minimum level of detail.
+The network $G_L$ is thus the initial focus of user interaction.
+The focus can be directed to other levels of the multilevel hierarchy, but a single level can be the current focus to which any interaction actions will apply.
+%to the representation at the level defined as the current focus.
+%There are multiple levels of representation entailed by the multilevel strategy.
+%This makes pertinent that one and only of the available levels is explicitly selected so that the transformations are performed on that level.
+Thus, there is always a user-selected level to which interface controls apply, such as link coloring or expanding super-nodes to reveal their predecessor nodes.
+As the reader may infer, the numbers of visible nodes and links in each level are not fixed throughout the navigation.
+The user can modify the focus level in the table to the right of the canvas, as shown in Figure~\ref{secondPhase}, which also shows the number of visible and total nodes and links at each level.
+The user may also interact with the table to modify node color and shape or link color at any layer/level, as well as toggle show/hide links with left/right clicks on the corresponding colored cell.
 
-\begin{figure}[!h]\centering
- \includegraphics[width=\textwidth]{secondPhase}
-  \caption{The elements in second step of the MlBiNetVis usage, dedicated to navigation on the network and transformation of the visualization.
-  The user may change the layout used when supernodes are opened and the predecessors are then shown on the screen.
-  Links may be shown or hided for all the levels by using the checkbutton.
-  The text area assists in the output of information requested or useful for navigation.
-  The canvas holds the node-link representations of the levels, which are transformed using the tools in the toolbar and the table in the right-side.
-  Such table also makes evident key information about the status of the visualization.
-  Further information is given in Section~\ref{use}.
+\begin{figure}\centering
+ \includegraphics[width=\textwidth]{secondPhase_}
+  \caption{Interface elements in BiNetVis in the visualization mode, where the user can navigate in the network and modify visualization settings.
+  S/he may change the layout used when super-nodes are expanded and their predecessors rendered.
+The checkbox to the left of each level may be used to show or hide the corresponding links at the level.
+  The text area displays user-requested information or information useful for navigation.
+  The canvas holds the node-link representation of the network, possibly depicting nodes at multiple levels of the multilevel hierarchy.
+The node-link representation may be modified using the toolbar controls and the controls the table on the right, which also shows key information on the status of the visualization.
+Here the current focus is on level 8 from a multilevel network with 12 hierarchical levels.
+  %Further information is given in Section~\ref{use}.
   }\label{secondPhase}
@@
 
-The toolbar holds buttons for further actions to be performed and are divided in groups.
-The first group is for changing how the nodes are visualized: size, proportionality to
-the number of neighbors or predecessors, reset proportionality, transparency, and rotation.
-The second group of buttons is for the links: thickness, transparency, proportionality to weight
+The toolbar on top of the canvas in Figure~\ref{secondPhase} holds further controls organized in groups according to their target interface element.
+The first group of controls affect graphical attributes of nodes: size mapped to degree or to number of predecessors, reset size, modify transparency and rotation.
+The second group affects link properties: thickness, transparency, proportionality to weight
 using thickness or transparency, and reset proportionality.
 
-The third group of buttons is more specific to the needs of the MlBiNetViz, and
-are selected to make mouse operations entail special processes and transformations of the visualization.
-The first button is selected for the interface to return core information from a clicked node:
-id, level, number of predecessors of previous level, id of successor (if node is not on the last level),
-number of neighbors (degree), sum of weight of links (strength).
-Other information the analysts report necessity of are easily implemented, but currently only these informations
-are returned to avoid unecessarily flooding the text area.
-The second tool button is seleted to move nodes: specific nodes are individually moved if clicked and dragged;
-a rectangular region is made available for moving all nodes therein if an empty region of the canvas
-is clicked and the mouse cursor dragged before releasing the button.
-The third tool button is selected for opening supernodes and exposing their predecessor nodes and links among them:
-a region is defined by the user by clicking on an empty space on the canvas and gragging the mouse before releasing,
-the supernodes therein are joined into a rectangular region that contains all of them,
-and their predecessor nodes (which are on a finer level of detail) are drawn inside such rectangular region.
-Open supernodes may be joined using the same tool, but it was found convenient to make available a
-fourth button to assist in this task: if this tool is selected, one may join two open supernodes by
-clicking on them in sequence.
-The last tool button in this group is selected to perform three operations on open supernodes:
-if an open supernode is clicked on and dragged, the rectangular shape is resized accordingly;
-if there is no drag, i.e. the super node is clicked on and released, the links from it to the other
-supernodes are attached to the next counterclockwise corner of its rectagular region.
-When the open supernode is resized, the layout of its child nodes is optimized for the drawn region.
-This is most useful when used in combination with the drag tool:
-on the finer level of detail, the nodes are repositioned into a more compact disposition.
-Then upon a (potentially minimal) supernode
-risize, the child nodes are made to fit the whole rectangular area.
-The last set of tools acts on the whole canvas:
-changing the color of the background, zoom, left-right/up-down pan,
-rotation of the entire network,
-and toggling between current zoom and pan status and the initial setting.
-The toolbar is contextualized in Figure~\ref{secondPhase}.
+The third group refers to controls more specific to how BiNetVis operates as a multilevel visualization. 
+%are selected to make mouse operations entail special processes and transformations of the visualization.
+The first control returns information relative to a selected (clicked) node:
+id, level, number of predecessors at the upper level, successor id (if applicable), degree, strength (sum of link weights).
+%Other information the analysts report necessity of are easily implemented, but currently only these items are returned to avoid unnecessarily flooding the text area.
+The second control allows moving nodes: specific nodes are moved if clicked and dragged; a rectangular region can be defined (clicking on an empty region in the canvas and dragging then releasing the mouse) to move all nodes within.
+The third control allows expanding super-nodes to expose their predecessor nodes and links.
+Again, the user specifies a rectangular region and the super-nodes within are bounded within the region and replaced with their predecessor nodes, which exist at the upper hierarchical level (thus at a greater level of detail).
+Thus, an expanded super-node is identified by a bounding rectangle that contains a view of its predecessor nodes and links.
+%are drawn inside this region.
+The same control can be employed to join expanded super-nodes, but it was found convenient to include a fourth control to assist in this task, which allows joining two expanded super-nodes by clicking on them in sequence.
+The final control in this group supports three operations on expanded super-nodes:
+if an expanded super-node is clicked on and dragged its rectangular shape is resized accordingly;
+if the super-node is clicked on and released (no dragging) its links  to the other super-nodes are shown attached to the next counterclockwise corner of the bounding rectangle.
+When the expanded super-node is resized the spatial layout of its child nodes is optimized to the modified drawing area.
+This is most useful in combination with the drag tool, to rearrange the nodes at the finer level of detail in a more compact disposition.
+Upon a super-node resize the child nodes are rearranged to fit the new rectangular area.
+The final set of controls act on the canvas: they allow
+changing the background color, zooming, left-right/up-down panning,
+rotating the entire network, and toggling between the current zoom and pan configuration and the initial settings.
+%The toolbar is contextualized in Figure~\ref{secondPhase}.
 
-There are many subleties on the orchestrated use of these tools.
-As a facilitated means to convey the navigation possibilities and
-procedures available, the user is invited to watch a demo video in~\cite{yvideo},
-and use the MlBiNetVis interface through the instance we made availabe~\cite{mbnvpage}.
-The more seasoned software developer may browse the code, upload a local instance
-of the software, and make changes to its source code~\cite{mbnvcode}.
+There are many subtleties on the orchestrated use of these controls.
+As a facilitated means to convey the navigation possibilities, the user is invited to watch a demo video~\footnote{\url{http://rfabbri.vicg.icmc.usp.br:3000/multilevel2/about}} %in~\cite{yvideo}
+and use the BiNetVis interface~\footnote{\url{http://rfabbri.vicg.icmc.usp.br:3000/multilevel2/topdown}}.
+Furthermore, a simple case study is described in Appendix B.
+%\cite{mbnvpage}.
+The more seasoned software developers may browse the code, upload a local instance of the software, and make changes to its source code~\footnote{ https://github.com/ttm/netText}.
+%\cite{mbnvcode}.
 
-\subsection{Software technologies employed within mlvis}
+\subsection{Underlying software technologies}
 % coarsening algorithms
@@
 % Nuxt/Vue.js for the frontend with heavy use of PIXI.js
-Most important parts of MlBiNetViz are written in a combination of JavaScript and Python:
-Vue.js (set up by Nuxt.js) is used to assist in the client (frontend),
-the server (backend) is a Flask Python server, used to perform
-specialized or heavy calculations.
-There is a secondery server, a FeatherJS, used to facilitate contact with the database
-and real-time multi-user interaction.
-The data is stored in a MongoDB database and ordinarily in the filesystem,
-while the multi-user interaction was deactivated for now to avoid unecessary complexity.
-The coarsening algorithms for bipartite networks are available in~\cite{bialgs}
-and are used by the Flask server.
+Most important components of BiNetVis are written in a combination of JavaScript and Python: it uses Vue.js (set up by Nuxt.js) in the frontend client, the backend is a Flask Python server, used to perform specialized or heavy calculations.
+A secondary server, a FeatherJS, is used to facilitate contact with the database and real-time multi-user interaction.
+The data is stored in a MongoDB database and ordinarily in the file system,
+while the multi-user interaction is deactivated for now to avoid unnecessary complexity.
+Multiple coarsening algorithms for bipartite networks available from a previous implementation~\cite{alan2} are accessed by the Flask server.
 The fast WebGL 2D rendering on the canvas is performed using Pixi.js.
 
-The rasterization of images is usually performed though triangles.
-Accordingly, in order to comply with the goal of enabling the visualization of large bipartite networks,
-the rendered network on the canvas contains only triangles (one for each of the nodes) and straight lines (for the links).
-The user may wish to further alleviate the computational cost by not showing the links.
-To emphasize the bipartite structure of the network, and for making the visualization more aesthetically
-compelling, the user may change the shape of the nodes in one (or both) of the layers to hexagons instead of triangles.
+%Scene rasterization is usually performed though triangles.
+In order to comply with the goal of handling large bipartite networks, the network geometry on the canvas uses only triangles as primitives (for nodes) and straight lines (for links).
+The user may wish to further alleviate the computational burden by not rendering the links.
+To emphasize the bipartite nature of the network and render the visualization more aesthetically compelling the user may modify the node shapes in one (or both) layers to hexagons, instead of triangles.
 Using these features, and within the technologies described, we have visualized networks
@@
 with links shown,
-even using ordinary machines (e.g. with 8GB RAM DDR3, an i7 processor of the first generations, and a 1GB GPU).
+even when running the system on ordinary machines, e.g. with 8GB RAM DDR3, a first generation i7 processor and a 1GB GPU.
 
-This is a very brief account of the styles and software resources used to implement
-MlBiNetVis. The interested reader is welcome to browse the software code and documentation
-and contact the authors~\cite{mbnvcode}.
+%This is a brief account of the resources implemented in BiNetVis, the interested reader is welcome to browse the software code and documentation and contact the authors.
+%~\cite{mbnvcode}.
 
@@
 % extension to multipartite (or heterogeneous) networks.
-We have described and made available the first visualization interface for bipartite networks
-assited by multilevel strategies.
-The resulting conceptual organization is very consistent with the Shneiderman's mantra,
-as exemplified the by the implemented software.
-The interface developed is efficiently managed through simple interactive operations,
-and very efficient in visualization large networks due to the use of multilevel strategies,
-the software technologies employed, and the computationally inexpensive design decisions regarding
-the rendering of the node-link diagrams.
-In summary, this work is a proof-of-concept in that it establishes as feasible the use
-of multilevel strategies for the visualization of large bipartite networks.
+We introduced a visualization interface for bipartite networks
+assisted by the multilevel method that admits a conceptual organization very consistent with the well-known visual information seeking mantra stated as \emph{overview first, zoom and filter, then details-on-demand}.
+As exemplified in the software description, it allows users to obtain an overview of the major topological structures in a network and then focus on relevant elements for which further details can be displayed.
+The combination of the multilevel strategy with suitable software technologies and computationally inexpensive design decisions regarding
+the rendering of node-link representations yields a visual interface manageable with simple interactive operations that can effectively display large-scale networks.
+In summary, this software is introduced as a proof-of-concept on the feasibility of employing multilevel strategies for using the familiar node-link views  to visualize large bipartite networks.
+The same underlying rationale is applicable to unipartite and to heterogeneous networks as long as the underlying multilevel methods are provided.
 
-In further steps, the navigation may be enhanced by the acute use of in-development coarsening algorithms.
-The visualization pathway is being envisioned as capable to assist the development of novel
-multilevel strategies, which should be verified.
-Also, there are multilevel strategies that involve the selection of nodes as pivors
-to guide the coarsening procedure, a process to which there is no visual interface,
-and to which MlBiNetVis should be adapted.
-Also, the interface is envioned to assist the analysis of coarsening results. As in the current version of MlBiNetVis,
-the analyst is able to scrutinize linking patterns and overall network structure though the node-link diagram,
-and requesting information, but this should be enhanced by the display of histograms and global measures of each level.
-On the analysis of the networked structures, the navigation involving metadata, such as from genes and proteines, or as
-documents and authors, should result in a visual analytics interface for interactive knowledge discovery.
-Finally, bipartite networks are naturally extended to heterogeneous (or multipartite) networks,
-and thus all the content of this article, and should require further research and development for the achievement
-of concrete derivatives.
+The proposed visualization solution could be incorporated into a visual analytics environment for large networked datasets, with added tools to enable data analytics.
+It does require further validation on practical analytical settings, as the  usage scenarios of interactive knowledge discovery are inherently complex.
+We believe that tuning this implementation into a tool to guide and inform developers of novel multilevel algorithms and applications would also be an interesting development.
+For instance, certain multilevel strategies require selecting pivot nodes to guide the coarsening procedure, a task that could benefit from an interactive visual interface.
+The same applies to developers who wish to compare the outcome of  multiple executions of coarsening algorithms, e.g., with different parameter settings.
 
-\appendix
+%In further steps, the navigation may be enhanced in multiple ways. 
+%by the acute use of in-development coarsening algorithms.
+%The system may be useful to assist the development of novel multilevel strategies, which should be verified.
+There are multilevel strategies that involve the selection of nodes as pivots to guide the coarsening procedure, a process to which there is no visual interface, and to which BiNetVis could be adapted.
+%The interface is also envisioned to assist the analysis of coarsening results.
+In the current version of BiNetVis the analyst is able to scrutinize linking patterns and overall network structure though the node-link diagram, and request information, but it should be enhanced, e.g., to display histograms and global measures relative to each level.
+%On the analysis of the networked structures, the navigation involving metadata, such as from genes and proteins, or from  documents and authors, should result in a visual analytics interface for interactive knowledge discovery.
+%Finally, bipartite networks are naturally extended to heterogeneous (or multipartite) networks, and thus all the content of this article, and should require further research and development for the achievement of concrete derivatives.
 
-\section{Choice and fine-tuning of the coarsening algorithm}\label{ap}
-The elements in Figure~\ref{initial}.
+\subsubsection*{Acknowledgements}
+The authors acknowledge the financial support of the S\~ao Paulo State Research Foundation (FAPESP grant 2017/05838-3) and the National Council for Scientific and Technological Development (CNPq, grant 301847/2017-7).
+The views expressed do not reflect the official policy or position of FAPESP or CNPq.
+
+%\appendix
+
+%\section{Choice and fine-tuning of the coarsening algorithm}\label{ap}
+%The elements in Figure~\ref{initial}.
 
@@
 % \end{thebibliography}
+
+\appendix
+
+\newpage
+\section*{Appendix A: Coarsening parameters}
+
+Table \ref{tab:parameters} lists the input parameters of the coarsening algorithms, which can be applied independently to one or to both network layers. 
+%to fit their corresponding parameters. 
+Three parameters are common to all algorithms, namely the maximum number of coarsening levels~\textit{max-levels}, a reduction factor~\textit{reduction} and a similarity function~\textit{similarity}, which returns a similarity score between a pair of vertices.
+Parameter ~\textit{reduction} is multiplied by the number of vertices to determine the maximum number of nodes matched, e.g.
+if $reduction = 0.5$ each coarsening iteration will (potentially) reduce the number of nodes by a factor of two, yielding a logarithmic decrease in network size along the process.
+%Similarity defines the distance between two vertices, in which a high similarity score represents a high degree of similarity. 
+A possible similarity function is the \emph{common neighbors} metric, which returns the number of common neighbors between the two nodes.
+The default coarsening algorithm in the system is $MLPb$, which requires four additional parameters: \textit{min-vertices}, which is the minimum number nodes expected in the coarsest network and two parameters that serve as stopping criteria, namely a \textit{maximum number of iterations} and a \textit{tolerance} threshold.
+The algorithm stops if either the maximum number of iterations has been reached or the number of label updates is bellow the defined tolerance.
+A final parameter, namely \textit{upper bound}, defines an upper limit on the weights of super-nodes, which is useful to create super-nodes more faithful to the original structures and to avoid super-node weights that deviate significantly from the average values.
+%that can be manipulated to create the coarsening networks.
+
+\begin{table}[!ht]
+    \centering
+	\caption{Parameters of the multilevel coarsening algorithms (see Section~\ref{smul}). The rightmost column reports the default values. }
+	%algorithms that use the respective parameter.}
+	\label{tab:parameters}
+ 	\renewcommand{\arraystretch}{1.3}
+ 	\renewcommand{\tabcolsep}{0.45cm}
+    \begin{tabular}{lrr}
+    	\textbf{Parameter} & \textbf{Domain} & \textbf{Default} \\
+    	\hline
+    	%matching & string & MLPb \\
+    	similarity & string & Common Neighbor \\
+        reduction & $(0,1] \subset \mathbb{R}_+$ & 0.1 \\
+        max-levels & $[0,n] \subset \mathbb{Z}_+$ & 5 \\
+        min-vertices & $[0,n-1] \subset \mathbb{Z}_+$ & 100 \\
+        upper-bound & $(0,n] \subset \mathbb{R}_+$ & 0.1 \\
+        iterations & $\subset \mathbb{Z}_+$ & 10 \\
+        tolerance & $\subset \mathbb{R}_+$ & 0.0001 \\
+    \end{tabular}
+\end{table}
+
+\null
+\vfill
+
+\section*{Appendix B: Case study}
+The interaction patterns between genes and proteins cast the basis of molecular biology and of disease pathogenesis~\cite{netD,mlCa}.
+Protein groups are defined by topological characteristics yield by the networks they entail,
+allowing for the isolation of functional and disease pathways~\cite{bara,shara}.
+Given these premises,
+a brief case study was performed considering the protein and
+n-reactome iterations network,
+which has 8788 proteins, 15433 iterations, and 41087 links.
+Figure~\ref{aml0} (a) depicts the original network.
+There are three communities that are more expressive and smaller communities.
+Figure~\ref{aml0} (b) illustrates a reduced representation of the network,
+with 200 nodes in each layer,
+which resulted from 3 contraction steps.
+The reduced network highlights the larger communities,
+mirroring predominant characteristics of the starting network.
+As in this example, in general,
+the multilevel strategy provides good insight about the original topology.
+
+\begin{figure}[!h]\centering
+    \subfloat[original]{{\includegraphics[width=.46\textwidth, height=4.3cm]{aor} }}%
+    \qquad
+  \subfloat[coarsened]{{\includegraphics[width=.46\textwidth, height=4.3cm]{aml} }}%
+    \caption{Graphical representations of a network with 24,211 nodes (8,788 proteins and 15,422 n-reactome iterations), and 41,087 links.
+    In (a) the original network is mapped visually, and in (b) the network is represented using only 200 nodes (100 nodes in each layer).
+    }%
+    \label{aml0}%
+\end{figure}
+
+For this case study, we focused in the smallest of the three communities,
+in order to visualize it separately,
+observe its connectivity patterns,
+and consequently the functional patterns
+of the nodes that belong to this community.
+Figures~\ref{aml1} (a), (b) and (c) depict the community in levels 2, 1 and 0 (original), respectively.
+The final result is isolated in Figure~\ref{amlf},
+were only the community is visualized, and only with uncoarsened nodes.
+Therein, it is evident that a group of more connected nodes exists, they are in the center of the image.
+The analist may wish, e.g., to target the genes among such nodes
+in further steps of the investigation,
+due to their central role in interacting with numerous proteins.
+Another possibility is to examine the less connected nodes in an attempt
+to understand their specific roles.
+
+\begin{figure}[!h]\centering
+    \subfloat[original]{{\includegraphics[width=.3\textwidth, height=3.3cm]{aml2} }}%
+  \subfloat[coarsened]{{\includegraphics[width=.3\textwidth, height=3.3cm]{aml3} }}%
+  \subfloat[coarsened]{{\includegraphics[width=.3\textwidth, height=3.3cm]{aml4_} }}%
+    \caption{Progressive visualization of the smallest community if Figure~\ref{aml0}. In (a), the nodes are uncoarsened to level 2.
+    In (b), the nodes are uncorsened to level 1.
+    In (c), the noeds are uncoarsened to level 0, i.e. to the level of the original, uncoarsened, network.}%
+    \label{aml1}%
+\end{figure}
+
+\begin{figure}\centering
+ \includegraphics[width=0.7\textwidth,height=7cm]{aml5_}
+  \caption{An isolated visualization of the community being analysed.
+  The researcher reached this visualization by on-demand transformations of the initial visualization, i.e. from a respresentation with only 200 nodes of a network with 24,211 nodes.}\label{amlf}
+\end{figure}
+
+As exempified,
+given the importance of the connectivity patterns,
+such as group structure,
+and given the effectiveness of the visualization assisted by multilevel strategies in bipartite networks,
+the method may assist research related to gene/protein networks.
+Thus, BiNetVis is potentially helpful in studying pathogenic variants,
+groups of genes associated with metabolic deficiency,
+or groups of proteins that can be targeted for therapeutic goals.
+
+\FloatBarrier
+\section*{Appendix C: The toolbar}
+One special element that enables the interactive visualization of BiNetVis is the toolbar. It is contextualized in Figures~\ref{fpage0} and~\ref{secondPhase},
+but in a scale which may not favor its inspection by the reader, specially in a printed version of this document.
+Figure~\ref{toolbar} is dedicated to the toolbar and its usage is described in Section~\ref{use}.
+
+\begin{figure}\centering
+ \includegraphics[width=\textwidth]{toolbar}
+  \caption{The toolbar, contextualized in Figures~\ref{fpage0} and~\ref{secondPhase}, shown independently.
+The four groups of buttons are for transformations of: nodes, links, special/multilevel, and whole network.
+The usage is described in Section~\ref{use}.}\label{toolbar}
+\end{figure}
+
 \end{document}

### 2019-06-12 paper.tex
arquivos todos usados até a submissão

@@
 ``unipartite'') networks~\cite{u1,u2,u3,u4,u5,u6,u7,u8,u9,u10}.
-We are aware of a single (yet unpublished) contribution in the context of bipartite networks, in which a visual metaphor is used to assist users executing multilevel methods~\cite{cintra2019}.
+We are aware of a single contribution (yet unpublished) in the context of bipartite networks, in which a visual metaphor is devised to assist users executing multilevel methods~\cite{cintra2019}.
 In general, solutions for node-link visualizations of large networks often rely on aggregation of nodes in clusters or communities to reduce visual clutter and create simpler representations amenable to user interaction~\cite{a1,a2,a3,a4,a5,a6}.
@@
 
-The coarsening phase creates a hierarchy of coarsened networks $G_l$, $l$ integer $ \in [0,L]$ where $G_0$ is the original network and $|V_i| \geq |V_{i+1}|$, resulting in a hierarchical representation of $G_0$ at decreasing levels of detail.
+\noindent The coarsening phase creates a hierarchy of coarsened networks $G_l$, $l$ integer $ \in [0,L]$ where $G_0$ is the original network and $|V_i| \geq |V_{i+1}|$, resulting in a hierarchical representation of $G_0$ at decreasing levels of detail.
 The process is carried out by two algorithms: \emph{matching}, which decides which nodes to merge,
@@
 
-The matching is followed by the contraction of the network into a coarser form.
+\noindent The matching is followed by the contraction of the network into a coarser form.
 Typically, the nodes matched are merged into a super-node weighted according to the number of of its composing nodes, and the links incident to such nodes are joined into super-links,
@@
 \begin{algorithm}
-  \qquad\qquad\qquad{\bf\emph{BiNetVis Algorithm}}\\
-  \vspace{0.3cm}
+  %\qquad\qquad\qquad{\bf\emph{BiNetVis Algorithm}}\\
+  %\vspace{0.3cm}
   \KwIn{\\
@@
   \caption{
-An algorithmic description of the multilevel method adapted for the visualization of bipartite networks, named the {\bf \emph{BiNetVis Algorithm}}.
+%An algorithmic description of the multilevel method adapted for the visualization of bipartite networks, named the {\bf \emph{BiNetVis Algorithm}}.
+{\bf \emph{BiNetVis}} algorithm for multilevel visualization of bipartite networks.
   The routines {\bf Matching$_b$} and {\bf Contracting$_b$} refer to any matching or contracting
@@
 
-Other operations are convenient to adequate the visual mapping to the current user needs, e.g., setting the network size, expanding super-nodes, exposing levels and highlighting topological features.
+\noindent Other operations are convenient to adequate the visual mapping to the current user needs, e.g., setting the network size, expanding super-nodes, exposing levels and highlighting topological features.
 These operations are further detailed in Section~\ref{sof}.
@@
   \caption{The BiNetVis interface: (1) drop-down menu for selecting or uploading a network for analysis;
-  (2) control widgets to select and parameterize the coarsening algorithm;  (3) the render button to display the network; (4) auxiliary widgets for selecting the network layout, show/hide links, and area to display user requested information on nodes or links; (5) interactive table holding information about the multilevel hierarchy levels and the bipartite network layers; can also be used to set the visualization on the canvas; (6) toolbar with controls for navigation of the network and fine-tuning the visualizationi, better inspected in Figure~\ref{toolbar} of Appendix C; (7) canvas with the node-link representation of the multilevel network.
+  (2) control widgets to select and parameterize the coarsening algorithm;  (3) the render button to display the network; (4) auxiliary widgets for selecting the network layout, show/hide links, and area to display user requested information on nodes or links; (5) interactive table holding information about the multilevel hierarchy levels and the bipartite network layers; can also be used to set the visualization on the canvas; (6) toolbar with controls for navigation of the network and fine-tuning the visualization (detailed in Figure~\ref{toolbar} of Appendix B); (7) canvas with the node-link representation of the multilevel network.
 Notice that nodes shown belong to four distinct hierarchical levels: level 3 (dark red/green), level 2 (pink/cyan), level 3 (navy blue/yellow) and level 2 (red/bright green).
@@
   The button at the top is for network uploading and selection.
-  In the gray box, the user chooses the coarsening algorithm and sets its parameters, as detailed in the Appendix A. % Appendix~\ref{ap}.
+  In the gray box, the user chooses the coarsening algorithm and sets its parameters (Appendix A details the parameters of coarsening algorithms). % Appendix~\ref{ap}.
   The blue button at the bottom is a drop-down menu where the user can select the node-link layout algorithm, where multiple alternatives are provided.
@@
  \includegraphics[width=\textwidth]{secondPhase_}
-  \caption{Interface elements in BiNetVis in the visualization mode, where the user can navigate in the network and modify visualization settings.
+  \caption{Interface elements of BiNetVis in the visualization mode, where the user can navigate in the network and modify visualization settings.
   S/he may change the layout used when super-nodes are expanded and their predecessors rendered.
@@
 
-There are many subtleties on the orchestrated use of these controls.
-As a facilitated means to convey the navigation possibilities, the user is invited to watch a demo video~\footnote{\url{http://rfabbri.vicg.icmc.usp.br:3000/multilevel2/about}} %in~\cite{yvideo}
-and use the BiNetVis interface~\footnote{\url{http://rfabbri.vicg.icmc.usp.br:3000/multilevel2/topdown}}.
-Furthermore, a simple case study is described in Appendix B.
+\noindent There are many subtleties on the orchestrated use of these controls.
+As a facilitated means to convey the navigation possibilities, the user is invited to watch a demo video\footnote{\url{http://rfabbri.vicg.icmc.usp.br:3000/multilevel2/about}} %in~\cite{yvideo}
+and use the BiNetVis interface\footnote{\url{http://rfabbri.vicg.icmc.usp.br:3000/multilevel2/topdown}}.
+Furthermore, a simple case study is described in Appendix A.
 %\cite{mbnvpage}.
-The more seasoned software developers may browse the code, upload a local instance of the software, and make changes to its source code~\footnote{ https://github.com/ttm/netText}.
+The more seasoned software developers may browse the code, upload a local instance of the software, and make changes to its source code\footnote{ https://github.com/ttm/netText}.
 %\cite{mbnvcode}.
@@
 % Nuxt/Vue.js for the frontend with heavy use of PIXI.js
-Most important components of BiNetVis are written in a combination of JavaScript and Python: it uses Vue.js (set up by Nuxt.js) in the frontend client, the backend is a Flask Python server, used to perform specialized or heavy calculations.
+Components of BiNetVis are mostly written in a combination of JavaScript and Python: it uses Vue.js (set up by Nuxt.js) in the front-end client, the back-end is a Flask Python server, used to perform specialized or heavy calculations.
 A secondary server, a FeatherJS, is used to facilitate contact with the database and real-time multi-user interaction.
@@
 The user may wish to further alleviate the computational burden by not rendering the links.
-To emphasize the bipartite nature of the network and render the visualization more aesthetically compelling the user may modify the node shapes in one (or both) layers to hexagons, instead of triangles.
+To emphasize the bipartite nature of the network and render the visualization more aesthetically compelling, the user may modify the node shapes in one (or both) layers to hexagons, instead of triangles.
 Using these features, and within the technologies described, we have visualized networks
@@
 \newpage
-\section*{Appendix A: Coarsening parameters}
+
+\section*{Appendix A: A case study}
+The interaction patterns between genes and proteins cast the basis of molecular biology and of disease pathogenesis~\cite{mlCa,netD}.
+Protein groups are defined by topological characteristics yield by the networks they entail,
+allowing for the isolation of functional and disease pathways~\cite{bara,shara}.
+Given these premises,
+a brief case study was performed considering the protein and
+n-reactome iterations network,
+which has 8,788 proteins, 15,433 iterations, and 41,087 links.
+Figure~\ref{aml0}(a) depicts the original network. One observes three more expressive communities and several smaller communities.
+Figure~\ref{aml0}(b) illustrates a coarsened representation with 100 nodes in each layer obtained after three contraction steps.
+The coarsened network emphasizes the larger communities,
+mirroring predominant characteristics of the input network.
+%As in this example, in general, the multilevel strategy provides good insight about the original topology.
+
+\begin{figure}[!h]\centering
+    \subfloat[original]{{\includegraphics[width=.46\textwidth, height=4.3cm]{aor} }}%
+    \qquad
+  \subfloat[coarsened]{{\includegraphics[width=.46\textwidth, height=4.3cm]{01} }}%
+    \caption{Graphical representations of a network with 24,211 nodes (8,788 proteins and 15,422 n-reactome iterations) and 41,087 links.
+    (a) shows the original network; (b) shows the coarsest network represented with 200 nodes only (100 nodes in each layer).
+    }%
+    \label{aml0}%
+\end{figure}
+
+\noindent For this case study, we focused on visualizing the smallest community in order to observe its connectivity and functional node patterns.
+Figures~\ref{aml1}(a), (b) and (c) depict the community observed in levels 2, 1 and 0 (original), respectively.
+The final result (Figure~\ref{aml1}(c)) is displayed in Figure~\ref{amlf},
+%were only the community is visualized, and only with uncoarsened nodes.
+where a group of densely connected nodes can be observed in the central area. An analyst may wish, e.g., to target the genes among such nodes
+in further steps of the investigation,
+due to their central role in interacting with numerous proteins.
+Another possibility is to examine the less connected nodes in an attempt
+to understand their specific roles. 
+%As exemplified, given the importance of the connectivity patterns, such as group structure, the method may assist research related to gene/protein networks.
+The example illustrates the potential usefulness of BiNetVis in studying pathogenic variants, groups of genes associated with metabolic deficiency,
+or groups of proteins that can be targeted for therapeutic or other purposes.
+
+\begin{figure}[!h]\centering
+    \subfloat[level 2]{{\includegraphics[width=.33\textwidth, height=3.5cm]{02} }}%
+  \subfloat[level 1]{{\includegraphics[width=.33\textwidth, height=3.5cm]{03} }}%
+  \subfloat[level 0]{{\includegraphics[width=.33\textwidth, height=3.5cm]{04} }}%
+    \caption{Progressive visualization of the smallest community in Figure~\ref{aml0}.
+In (a) the community is shown at the coarsest level; in (b) it has been been uncoarsened to the next level; and in (c) it has been uncoarsened to the final level, as in the original network.}%
+    \label{aml1}%
+\end{figure}
+
+\begin{figure}\centering
+ \includegraphics[width=0.9\textwidth]{05}
+  \caption{An isolated visualization of the community under analysis.
+  The researcher reached this visualization by on-demand transformations of the initial visualization of the original network (24,211 nodes) coarsened to a representation with only 200 nodes.}\label{amlf}
+\end{figure}
+
+\FloatBarrier
+\section*{Appendix B: The toolbar}
+The interface element that enables the interactive visualization in BiNetVis is the toolbar. It was contextualized in Figures~\ref{fpage0} and~\ref{secondPhase} and in Section~\ref{use},
+but in a scale which may not favor its inspection by the reader, specially in a printed version of this document.
+Figure~\ref{toolbar} details the toolbar.
+
+\begin{figure}\centering
+ \includegraphics[width=\textwidth]{toolbar}
+  \caption{The toolbar, contextualized in Figures~\ref{fpage0} and~\ref{secondPhase}, shown independently.
+The four groups of buttons are for specifying transformations applied to nodes, links, special/multilevel, and whole network.
+The usage has been described in Section~\ref{use}.}\label{toolbar}
+\end{figure}
+
+\section*{Appendix C: Coarsening parameters}
 
@@
 A possible similarity function is the \emph{common neighbors} metric, which returns the number of common neighbors between the two nodes.
-The default coarsening algorithm in the system is $MLPb$, which requires four additional parameters: \textit{min-vertices}, which is the minimum number nodes expected in the coarsest network and two parameters that serve as stopping criteria, namely a \textit{maximum number of iterations} and a \textit{tolerance} threshold.
-The algorithm stops if either the maximum number of iterations has been reached or the number of label updates is bellow the defined tolerance.
-A final parameter, namely \textit{upper bound}, defines an upper limit on the weights of super-nodes, which is useful to create super-nodes more faithful to the original structures and to avoid super-node weights that deviate significantly from the average values.
+The default coarsening algorithm in the system is $MLPb$, which requires four additional parameters.
+Parameter \textit{min-vertices} informs the minimum number nodes expected in the coarsest network.
+Parameters \textit{max-levels} and  \textit{tolerance} define stopping criteria: the algorithm stops if either the maximum number of iterations has been reached or the number of label updates is bellow the tolerance threshold.
+A final parameter, namely \textit{upper bound}, defines an upper limit on the weights of super-nodes, which is useful to preserve the original structures when creating super-nodes and to avoid super-nodes with weights that deviate significantly from the average.
 %that can be manipulated to create the coarsening networks.
@@
 
-\section*{Appendix B: Case study}
-The interaction patterns between genes and proteins cast the basis of molecular biology and of disease pathogenesis~\cite{netD,mlCa}.
-Protein groups are defined by topological characteristics yield by the networks they entail,
-allowing for the isolation of functional and disease pathways~\cite{bara,shara}.
-Given these premises,
-a brief case study was performed considering the protein and
-n-reactome iterations network,
-which has 8788 proteins, 15433 iterations, and 41087 links.
-Figure~\ref{aml0} (a) depicts the original network.
-There are three communities that are more expressive and smaller communities.
-Figure~\ref{aml0} (b) illustrates a reduced representation of the network,
-with 200 nodes in each layer,
-which resulted from 3 contraction steps.
-The reduced network highlights the larger communities,
-mirroring predominant characteristics of the starting network.
-As in this example, in general,
-the multilevel strategy provides good insight about the original topology.
-
-\begin{figure}[!h]\centering
-    \subfloat[original]{{\includegraphics[width=.46\textwidth, height=4.3cm]{aor} }}%
-    \qquad
-  \subfloat[coarsened]{{\includegraphics[width=.46\textwidth, height=4.3cm]{aml} }}%
-    \caption{Graphical representations of a network with 24,211 nodes (8,788 proteins and 15,422 n-reactome iterations), and 41,087 links.
-    In (a) the original network is mapped visually, and in (b) the network is represented using only 200 nodes (100 nodes in each layer).
-    }%
-    \label{aml0}%
-\end{figure}
-
-For this case study, we focused in the smallest of the three communities,
-in order to visualize it separately,
-observe its connectivity patterns,
-and consequently the functional patterns
-of the nodes that belong to this community.
-Figures~\ref{aml1} (a), (b) and (c) depict the community in levels 2, 1 and 0 (original), respectively.
-The final result is isolated in Figure~\ref{amlf},
-were only the community is visualized, and only with uncoarsened nodes.
-Therein, it is evident that a group of more connected nodes exists, they are in the center of the image.
-The analist may wish, e.g., to target the genes among such nodes
-in further steps of the investigation,
-due to their central role in interacting with numerous proteins.
-Another possibility is to examine the less connected nodes in an attempt
-to understand their specific roles.
-
-\begin{figure}[!h]\centering
-    \subfloat[original]{{\includegraphics[width=.3\textwidth, height=3.3cm]{aml2} }}%
-  \subfloat[coarsened]{{\includegraphics[width=.3\textwidth, height=3.3cm]{aml3} }}%
-  \subfloat[coarsened]{{\includegraphics[width=.3\textwidth, height=3.3cm]{aml4_} }}%
-    \caption{Progressive visualization of the smallest community if Figure~\ref{aml0}. In (a), the nodes are uncoarsened to level 2.
-    In (b), the nodes are uncorsened to level 1.
-    In (c), the noeds are uncoarsened to level 0, i.e. to the level of the original, uncoarsened, network.}%
-    \label{aml1}%
-\end{figure}
-
-\begin{figure}\centering
- \includegraphics[width=0.7\textwidth,height=7cm]{aml5_}
-  \caption{An isolated visualization of the community being analysed.
-  The researcher reached this visualization by on-demand transformations of the initial visualization, i.e. from a respresentation with only 200 nodes of a network with 24,211 nodes.}\label{amlf}
-\end{figure}
-
-As exempified,
-given the importance of the connectivity patterns,
-such as group structure,
-and given the effectiveness of the visualization assisted by multilevel strategies in bipartite networks,
-the method may assist research related to gene/protein networks.
-Thus, BiNetVis is potentially helpful in studying pathogenic variants,
-groups of genes associated with metabolic deficiency,
-or groups of proteins that can be targeted for therapeutic goals.
-
-\FloatBarrier
-\section*{Appendix C: The toolbar}
-One special element that enables the interactive visualization of BiNetVis is the toolbar. It is contextualized in Figures~\ref{fpage0} and~\ref{secondPhase},
-but in a scale which may not favor its inspection by the reader, specially in a printed version of this document.
-Figure~\ref{toolbar} is dedicated to the toolbar and its usage is described in Section~\ref{use}.
-
-\begin{figure}\centering
- \includegraphics[width=\textwidth]{toolbar}
-  \caption{The toolbar, contextualized in Figures~\ref{fpage0} and~\ref{secondPhase}, shown independently.
-The four groups of buttons are for transformations of: nodes, links, special/multilevel, and whole network.
-The usage is described in Section~\ref{use}.}\label{toolbar}
-\end{figure}
-
 \end{document}
